---
title: "GB_Mmul_10_snRNAseq_cellranger_ensembl_5_16_25"
output: html_document
date: "2025-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Set working directory
```{r}
setwd("/home/gb1-umw")
getwd()
```


# keyboard shortcut to add a code chunk in Rmd: command + option + i
# keyboard shortcut run current code chunk: command + shift + return 
# Can highlight multiple lines and run command + shift + return
# Delete chunk by highlighting and deleting it


# IN SUMMARY
# 18 SAMPLES/LIBRARIES mapped with cellranger count ENSEMBL files Macaca_mulatta.Mmul_10.dna.toplevel.fa & filtered_Macaca_mulatta.Mmul_10.113.gtf in ~/macaque_snRNAseq
# cellranger_count_1-MI1_230808_S3_run_5_15_25
# cellranger_count_1-MI3_V1_231103_S4_run_5_2_25
# cellranger_count_2-MI1_230808_S4_run_5_15_25
# cellranger_count_2-MI3_V1_231103_S5_run_5_2_25
# cellranger_count_3-C_MI2_230608_S1_run_5_15_25
# cellranger_count_3-MI1_V1_230809_S1_run_5_2_25
# cellranger_count_3-MI3_V1_231103_S6_run_5_2_25
# cellranger_count_4-D_MI2_230608_S2_run_5_15_25
# cellranger_count_4-MI1_230808_S5_run_5_15_25
# cellranger_count_4-MI3_V1_231103_S7_run_5_2_25
# cellranger_count_5-MI1_230808_S6_run_5_15_25
# cellranger_count_5-MI3_V1_231103_CAT_S1_run_5_13_25
# cellranger_count_6-MI1_V1_230809_S2_run_5_2_25
# cellranger_count_6-MI3_V1_231103_CAT_S2_run_5_13_25
# cellranger_count_7-MI1_230808_S7_run_5_15_25
# cellranger_count_7-MI3_V1_231103_CAT_S3_run_5_13_25
# cellranger_count_8-MI1_V1_230809_S3_run_5_2_25
# cellranger_count_8-MI3_V1_231103_CAT_S4_run_5_13_25


#8 reps from MI1
#2 reps from MI2
#8 reps from MI3



Load libraries (only run chunk if starting new session on cluster OOD). All packages loaded when logging onto RStudio server
```{r echo=FALSE}

suppressPackageStartupMessages(library(hdf5r)) #use to read in raw count matrix output from cell ranger
suppressPackageStartupMessages(library(Seurat)) #perform scRNA-seq processing & analysis
suppressPackageStartupMessages(library(SeuratDisk)) #SeuratDisk package is required to load the reference dataset in signac multiomocs tutorial
suppressPackageStartupMessages(library(Signac)) #perform snATACseq processing and analysis (also used for multiomocs analaysis)
suppressPackageStartupMessages(library(patchwork)) #makes plot creation in R faster & more powerful/efficient (works with ggplot2)
suppressPackageStartupMessages(library(ggplot2)) #create plots
suppressPackageStartupMessages(library(ggrepel)) #Provides text and label geoms for 'ggplot2' that help to avoid overlapping text labels
suppressPackageStartupMessages(library(tidyverse)) #used to clean & manipulate data
suppressPackageStartupMessages(library(parallel))  #used to parallelize R code so don't have to run 1 line at a time. Same with foreach, future, and doParallel
suppressPackageStartupMessages(library(foreach)) #used to parallelize R code
suppressPackageStartupMessages(library(future)) #used to parallelize R code
suppressPackageStartupMessages(library(doParallel)) #used to parallelize R code
suppressPackageStartupMessages(library(devtools)) #provides R functions that simplify and expedite common tasks in R code
suppressPackageStartupMessages(library(remotes)) #R installation from remote repositories from GitHub, etc
suppressPackageStartupMessages(library(GenomicRanges)) #used for storing and manipulating genomic intervals and variables defined along a genome reference/annotation
suppressPackageStartupMessages(library(rtracklayer)) #used for interacting with multiple genome browsers (currently UCSC built-in) and manipulating annotation tracks in various formats (GFF, BED, bedGraph, BED15, WIG, BigWig and 2bit)
suppressPackageStartupMessages(library(txdbmaker)) # set of tools for making TxDb objects from genomic annotations from various sources (e.g. UCSC, Ensembl, and GFF files). These tools allow the user to download the genomic locations of transcripts, exons, and CDS, for a given assembly, and to import them in a TxDb object. TxDb objects are implemented in the GenomicFeatures package, together with flexible methods for extracting the desired features in convenient formats.
suppressPackageStartupMessages(library(BSgenome)) # R package for (BS = biostrings) that's used to load in genomes (FASTA) from model organisms (human, mouse, macaque, etc)
# suppressPackageStartupMessages(library(BSgenome.Mfascicularis.NCBI.5.0)) # specifically load Macaca fascicularis FASTA from NCBI version 5.0. Similar to how signac tutorial has installed human FASTA
# suppressPackageStartupMessages(library(BSgenome.Mfascicularis.NCBI.6.0)) # specifically load Macaca fascicularis FASTA from NCBI version 6.0. Similar to how signac tutorial has installed human FASTA
# suppressPackageStartupMessages(library(SoupX)) #Identify ambient RNA contamination in single cell/nuclear data
suppressPackageStartupMessages(library(flexmix)) #miQC algorithm has been adapted for use in Seurat through installation of flexmix only
suppressPackageStartupMessages(library(R.utils)) #dependency for SeuratWrappers package
suppressPackageStartupMessages(library(SeuratWrappers)) #dependency for miQC to run on seurat objects
suppressPackageStartupMessages(library(miQC)) #Identify damaged nuclei
# suppressPackageStartupMessages(library(DropletQC)) #Identifies empty droplets in single cell/nuclear data
suppressPackageStartupMessages(library(DoubletFinder)) #Identifies doublets in single cell/nuclear data
suppressPackageStartupMessages(library(Rsamtools)) # Provides FaFile support to try with RegionStats function since can't modify BSgenome object for some reason
suppressPackageStartupMessages(library(ensembldb)) #Utilities to create and use Ensembl-based annotation databases
suppressPackageStartupMessages(library(AnnotationHub)) #fetch additional annotation resources from ensembl
suppressPackageStartupMessages(library(DESeq2)) #perform differential gene expression in Seurat FindMarkers function
suppressPackageStartupMessages(library(EnhancedVolcano)) #make fancier volcano plots!
suppressPackageStartupMessages(library(Matrix)) #matrix manipulation
suppressPackageStartupMessages(library(presto)) #implements a faster Wilcoxon test for FindMarkers function
suppressPackageStartupMessages(library(svglite)) #import images into adobe illustrator
suppressPackageStartupMessages(library(scCustomize)) #custom vizuals/functions for scRNAseq data

#use to keep output reproducible!
set.seed(6829) 

```


Run line from future package to run things in parallel (only run chunk if starting new session on cluster OOD)
```{r}
# from the future R package, which is used to enable parallel processing
# 20 parallel R sessions (or workers) for running tasks
# don't exceed workers by number of cores on OOD (32)
plan("multisession", workers = 20)
```


5/19/25
Read in filtered_feature_bc_matrix.h5 from each sample outs folder (18 samples total)
```{r}

# r objects can only have _ (underscore) and not dash - 
# can't start with a number
# this time leave out the basespace sample identifier (S3, S4, ETC because there are multiple of same ones depending on run)

# NOTE that rows in sparse matrix contains mmul10 gene id's (ex:ENSMMUG00000023296) as well as gene names (ZNF692)
# need to find a way to convert all rows to gene id's if possible? if id is in ensembl GTF then it doesn't have a gene name listed

# sample 1
sample_1MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_1-MI1_230808_S3_run_5_15_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 12337 sparse Matrix of class "dgCMatrix" - 26,530 genes & 12,337 nuclei

sample_1MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_1-MI3_V1_231103_S4_run_5_2_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9443 sparse Matrix of class "dgCMatrix"
  

# sample 2 
sample_2MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_2-MI1_230808_S4_run_5_15_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 11670 sparse Matrix of class "dgCMatrix"


sample_2MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_2-MI3_V1_231103_S5_run_5_2_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9581 sparse Matrix of class "dgCMatrix"

 
 # sample 3 
sample_3C_MI2_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_3-C_MI2_230608_S1_run_5_15_25/outs/filtered_feature_bc_matrix.h5")
# 26530 x 9976 sparse Matrix of class "dgCMatrix"

  
sample_3MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_3-MI1_V1_230809_S1_run_5_2_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 10609 sparse Matrix of class "dgCMatrix"
  
sample_3MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_3-MI3_V1_231103_S6_run_5_2_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9718 sparse Matrix of class "dgCMatrix"



  
# sample 4
sample_4D_MI2_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_4-D_MI2_230608_S2_run_5_15_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9651 sparse Matrix of class "dgCMatrix"
  
sample_4MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_4-MI1_230808_S5_run_5_15_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 7228 sparse Matrix of class "dgCMatrix"


sample_4MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_4-MI3_V1_231103_S7_run_5_2_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 9944 sparse Matrix of class "dgCMatrix"

  
 # sample 5 
sample_5MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_5-MI1_230808_S6_run_5_15_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9099 sparse Matrix of class "dgCMatrix"

sample_5MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_5-MI3_V1_231103_CAT_S1_run_5_13_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 9836 sparse Matrix of class "dgCMatrix"
 
  


# sample 6
sample_6MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_6-MI1_V1_230809_S2_run_5_2_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 11488 sparse Matrix of class "dgCMatrix"


sample_6MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_6-MI3_V1_231103_CAT_S2_run_5_13_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 8726 sparse Matrix of class "dgCMatrix"


  
# sample 7
sample_7MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_7-MI1_230808_S7_run_5_15_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 9664 sparse Matrix of class "dgCMatrix"

sample_7MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_7-MI3_V1_231103_CAT_S3_run_5_13_25/outs/filtered_feature_bc_matrix.h5") 
# 26530 x 9326 sparse Matrix of class "dgCMatrix"




# sample 8
sample_8MI1_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_8-MI1_V1_230809_S3_run_5_2_25/outs/filtered_feature_bc_matrix.h5")  
# 26530 x 9824 sparse Matrix of class "dgCMatrix"


sample_8MI3_matrix<-Read10X_h5("~/macaque_snRNAseq/cellranger_count_8-MI3_V1_231103_CAT_S4_run_5_13_25/outs/filtered_feature_bc_matrix.h5")  
#26530 x 12422 sparse Matrix of class "dgCMatrix"
 
  

```


Confirm each matrix is read in (DONE)
```{r}

sample_1MI1_matrix
sample_1MI3_matrix
sample_2MI1_matrix
sample_2MI3_matrix
sample_3C_MI2_matrix
sample_3MI1_matrix
sample_3MI3_matrix
sample_4D_MI2_matrix
sample_4MI1_matrix
sample_4MI3_matrix
sample_5MI1_matrix
sample_5MI3_matrix
sample_6MI1_matrix
sample_6MI3_matrix
sample_7MI1_matrix
sample_7MI3_matrix
sample_8MI1_matrix
sample_8MI3_matrix

```
# NOT running soupX and dropletQC (redundant with cellranger count pipeline)



Refer to Seurat tutorial https://satijalab.org/seurat/articles/pbmc3k_tutorial.html

Create seurat objects of all 18 libraries
so = seurat object
IF reading in rds seurat objects doesn't work, need to run chunk again and re-run subsequent chunks
```{r}

# a gene will only be retained in the Seurat object if it is detected in at least 3 cells.
# a cell will only be retained in the Seurat object if it expresses at least 200 genes.

sample_1MI1_so<-CreateSeuratObject(counts = sample_1MI1_matrix, min.cells = 3, min.features = 200)

sample_1MI3_so<-CreateSeuratObject(counts = sample_1MI3_matrix, min.cells = 3, min.features = 200)

sample_2MI1_so<-CreateSeuratObject(counts = sample_2MI1_matrix, min.cells = 3, min.features = 200)

sample_2MI3_so<-CreateSeuratObject(counts = sample_2MI3_matrix, min.cells = 3, min.features = 200)

sample_3C_MI2_so<-CreateSeuratObject(counts = sample_3C_MI2_matrix, min.cells = 3, min.features = 200)

sample_3MI1_so<-CreateSeuratObject(counts = sample_3MI1_matrix, min.cells = 3, min.features = 200)

sample_3MI3_so<-CreateSeuratObject(counts = sample_3MI3_matrix, min.cells = 3, min.features = 200)

sample_4D_MI2_so<-CreateSeuratObject(counts = sample_4D_MI2_matrix, min.cells = 3, min.features = 200)

sample_4MI1_so<-CreateSeuratObject(counts = sample_4MI1_matrix, min.cells = 3, min.features = 200)

sample_4MI3_so<-CreateSeuratObject(counts = sample_4MI3_matrix, min.cells = 3, min.features = 200)

sample_5MI1_so<-CreateSeuratObject(counts = sample_5MI1_matrix, min.cells = 3, min.features = 200)

sample_5MI3_so<-CreateSeuratObject(counts = sample_5MI3_matrix, min.cells = 3, min.features = 200)

sample_6MI1_so<-CreateSeuratObject(counts = sample_6MI1_matrix, min.cells = 3, min.features = 200)

sample_6MI3_so<-CreateSeuratObject(counts = sample_6MI3_matrix, min.cells = 3, min.features = 200)

sample_7MI1_so<-CreateSeuratObject(counts = sample_7MI1_matrix, min.cells = 3, min.features = 200)

sample_7MI3_so<-CreateSeuratObject(counts = sample_7MI3_matrix, min.cells = 3, min.features = 200)

sample_8MI1_so<-CreateSeuratObject(counts = sample_8MI1_matrix, min.cells = 3, min.features = 200)

sample_8MI3_so<-CreateSeuratObject(counts = sample_8MI3_matrix, min.cells = 3, min.features = 200)


```


Confirm each sample seurat object is read in
```{r}

sample_1MI1_so
sample_1MI3_so
sample_2MI1_so
sample_2MI3_so
sample_3C_MI2_so
sample_3MI1_so
sample_3MI3_so
sample_4D_MI2_so
sample_4MI1_so
sample_4MI3_so
sample_5MI1_so
sample_5MI3_so
sample_6MI1_so
sample_6MI3_so
sample_7MI1_so
sample_7MI3_so
sample_8MI1_so
sample_8MI3_so


```





5/20/25
View each Seurat object in console
```{r}

# > sample_1MI1_so
# An object of class Seurat 
# 21853 features across 12337 samples within 1 assay 
# Active assay: RNA (21853 features, 0 variable features)
#  1 layer present: counts
# > sample_1MI3_so
# An object of class Seurat 
# 20876 features across 9443 samples within 1 assay 
# Active assay: RNA (20876 features, 0 variable features)
#  1 layer present: counts
# > sample_2MI1_so
# An object of class Seurat 
# 21316 features across 11670 samples within 1 assay 
# Active assay: RNA (21316 features, 0 variable features)
#  1 layer present: counts
# > sample_2MI3_so
# An object of class Seurat 
# 20654 features across 9581 samples within 1 assay 
# Active assay: RNA (20654 features, 0 variable features)
#  1 layer present: counts
# > sample_3C_MI2_so
# An object of class Seurat 
# 20505 features across 9976 samples within 1 assay 
# Active assay: RNA (20505 features, 0 variable features)
#  1 layer present: counts
# > sample_3MI1_so
# An object of class Seurat 
# 21597 features across 10609 samples within 1 assay 
# Active assay: RNA (21597 features, 0 variable features)
#  1 layer present: counts
# > sample_3MI3_so
# An object of class Seurat 
# 20707 features across 9718 samples within 1 assay 
# Active assay: RNA (20707 features, 0 variable features)
#  1 layer present: counts
# > sample_4D_MI2_so
# An object of class Seurat 
# 20812 features across 9651 samples within 1 assay 
# Active assay: RNA (20812 features, 0 variable features)
#  1 layer present: counts
# > sample_4MI1_so
# An object of class Seurat 
# 21291 features across 7228 samples within 1 assay 
# Active assay: RNA (21291 features, 0 variable features)
#  1 layer present: counts
# > sample_4MI3_so
# An object of class Seurat 
# 20708 features across 9944 samples within 1 assay 
# Active assay: RNA (20708 features, 0 variable features)
#  1 layer present: counts
# > sample_5MI1_so
# An object of class Seurat 
# 21385 features across 9099 samples within 1 assay 
# Active assay: RNA (21385 features, 0 variable features)
#  1 layer present: counts
# > sample_5MI3_so
# An object of class Seurat 
# 21453 features across 9836 samples within 1 assay 
# Active assay: RNA (21453 features, 0 variable features)
#  1 layer present: counts
# > sample_6MI1_so
# An object of class Seurat 
# 20671 features across 11488 samples within 1 assay 
# Active assay: RNA (20671 features, 0 variable features)
#  1 layer present: counts
# > sample_6MI3_so
# An object of class Seurat 
# 21160 features across 8726 samples within 1 assay 
# Active assay: RNA (21160 features, 0 variable features)
#  1 layer present: counts
# > sample_7MI1_so
# An object of class Seurat 
# 21231 features across 9664 samples within 1 assay 
# Active assay: RNA (21231 features, 0 variable features)
#  1 layer present: counts
# > sample_7MI3_so
# An object of class Seurat 
# 21313 features across 9326 samples within 1 assay 
# Active assay: RNA (21313 features, 0 variable features)
#  1 layer present: counts
# > sample_8MI1_so
# An object of class Seurat 
# 20682 features across 9824 samples within 1 assay 
# Active assay: RNA (20682 features, 0 variable features)
#  1 layer present: counts
# > sample_8MI3_so
# An object of class Seurat 
# 21794 features across 12422 samples within 1 assay 
# Active assay: RNA (21794 features, 0 variable features)
#  1 layer present: counts


# TOTAL NUCLEI after creating seurat objects
# > 12337+9443+11670+9581+9976+10609+9718+9651+7228+9944+9099+9836+11488+8726+9664+9326+9824+12422
# [1] 180,542

```


View seurat object metadata
View(sample_1MI1_so@meta.data)

Save objects as rds just in case in cluster ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files  (DONE)
```{r}
# # Load necessary package (if not already loaded)
# library(Seurat)
# 
# # Output directory
# outdir <- "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files"
# dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
# 
# # List of Seurat object variable names
# sample_names <- c(
#   "sample_1MI1_so", "sample_1MI3_so", "sample_2MI1_so", "sample_2MI3_so",
#   "sample_3C_MI2_so", "sample_3MI1_so", "sample_3MI3_so", "sample_4D_MI2_so",
#   "sample_4MI1_so", "sample_4MI3_so", "sample_5MI1_so", "sample_5MI3_so",
#   "sample_6MI1_so", "sample_6MI3_so", "sample_7MI1_so", "sample_7MI3_so",
#   "sample_8MI1_so", "sample_8MI3_so"
# )
# 
# # since alot of samples, save each Seurat object in for loop and attach date
# for (s in sample_names) {
#   obj <- get(s)
# 
#   # Save with date in filename
#   saveRDS(obj, file = file.path(outdir, paste0(s, "_5_20_25.rds")))
# }


```



SKIP (DOESN'T WORK): Read in RDS files (doesn't re-read in as seurat objects)
```{r}

# indir <- "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files"
# rds_files <- list.files(indir, pattern = "\\.rds$", full.names = TRUE)
# 
# for (file in rds_files) {
#   # Read a single RDS file
#   obj <- readRDS(file)
# 
#   # Get the object name without .rds extension
#   obj_name <- gsub("\\.rds$", "", basename(file))
# 
#   # Optional: Print or do something with the object
#   cat("Loaded:", obj_name, "\n")
#   print(dim(obj))  # prints genes x cells
# 
#   # Optional: assign to environment (not required for line-by-line processing)
#   # assign(obj_name, obj)
# }


```


WORKS TO READ IN SEURAT OBJECTS: Try2 (substitute the _5_20_25 with nothing/remove it to reload in the sample seurat objects!!!!!)
```{r}

# gsub() is an R function that finds and replaces text in strings using regular expressions (regex).

# indir <- "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files"
# rds_files <- list.files(indir, pattern = "\\.rds$", full.names = TRUE)
for (file in rds_files) {
  obj <- readRDS(file)
  obj_name <- gsub("_5_20_25\\.rds$", "", basename(file))  # Remove suffix
  cat("Loaded:", obj_name, "\n")
  print(dim(obj))
  assign(obj_name, obj)
}

sample_names <- gsub("_5_20_25\\.rds$", "", basename(rds_files))

# reads in rds objects I made BEFORE running analysis steps. Need to re-run below chunks anyways but now know how to read in Seurat objects

```



Refer to GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd line 524
After making seurat objects for 7 samples, calculated percent.mt for each before running miQC.

Here calculate percent.mt & percent.rb

Calculate percent.mt
```{r}

# sample_1MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_1MI1_so, pattern = "^MT.")
# sample_1MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_1MI3_so, pattern = "^MT.")
# sample_2MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_2MI1_so, pattern = "^MT.")
# sample_2MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_2MI3_so, pattern = "^MT.")
# sample_3C_MI2_so[["percent.mt"]]<-PercentageFeatureSet(sample_3C_MI2_so, pattern = "^MT.")
# sample_3MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_3MI1_so, pattern = "^MT.")
# sample_3MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_3MI3_so, pattern = "^MT.")
# sample_4D_MI2_so[["percent.mt"]]<-PercentageFeatureSet(sample_4D_MI2_so, pattern = "^MT.")
# sample_4MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_4MI1_so, pattern = "^MT.")
# sample_4MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_4MI3_so, pattern = "^MT.")
# sample_5MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_5MI1_so, pattern = "^MT.")
# sample_5MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_5MI3_so, pattern = "^MT.")
# sample_6MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_6MI1_so, pattern = "^MT.")
# sample_6MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_6MI3_so, pattern = "^MT.")
# sample_7MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_7MI1_so, pattern = "^MT.")
# sample_7MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_7MI3_so, pattern = "^MT.")
# sample_8MI1_so[["percent.mt"]]<-PercentageFeatureSet(sample_8MI1_so, pattern = "^MT.")
# sample_8MI3_so[["percent.mt"]]<-PercentageFeatureSet(sample_8MI3_so, pattern = "^MT.")


```


Better way to calculate percent.mt for 18 samples. Also add percent.rb (re-run 5/29/25)
```{r}

# List of seurat object names
sample_names <- c(
  "sample_1MI1_so", "sample_1MI3_so", "sample_2MI1_so", "sample_2MI3_so",
  "sample_3C_MI2_so", "sample_3MI1_so", "sample_3MI3_so", "sample_4D_MI2_so",
  "sample_4MI1_so", "sample_4MI3_so", "sample_5MI1_so", "sample_5MI3_so",
  "sample_6MI1_so", "sample_6MI3_so", "sample_7MI1_so", "sample_7MI3_so",
  "sample_8MI1_so", "sample_8MI3_so"
)

# Loop through each object, calculate percent.mt & percent.rb, and update the object
for (s in sample_names) {
  obj <- get(s)
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT.")
  obj[["percent.rb"]] <- PercentageFeatureSet(obj, pattern = "^RP[SL]")
  assign(s, obj)
}

# check metadata to see if column added
# View(sample_1MI1_so@meta.data)


```


SKIP
Calculate percent.rb
```{r}
# Loop through each object, calculate percent.rb, and update the object
# for (s in sample_names) {
#   obj <- get(s)
#   obj[["percent.rb"]] <- PercentageFeatureSet(obj, pattern = "^RP[SL]")
#   assign(s, obj)
# }

# check metadata to see if column added
# View(sample_1MI1_so@meta.data)

```



Loop through samples to run miQC (re-run 5/29/25)
```{r}

# all packages loaded from beginning script


# Loop through each object, calculate percent.rb, and update the object
for (s in sample_names) {
  obj <- get(s)
  obj<-RunMiQC(obj, percent.mt = "percent.mt", nFeature_RNA = "nFeature_RNA", posterior.cutoff = 0.75, model.slot = "flexmix_model")
  assign(s, obj)
}

# check metadata to see if column added to all samples (YES!)
# View(sample_1MI1_so@meta.data)
# View(sample_7MI3_so@meta.data)



```

Move on to subsetting the cells where you only include the nuclei that says keep in the miQC.keep column
FIRST CHECK how many cells should be kept and how many should be discarded (re-run 5/29/25)
```{r}

for (s in sample_names) {
  obj <- get(s)
  cat("===", s, "===\n")
  print(table(obj@meta.data$miQC.keep))
}



# === sample_1MI1_so ===
# 
# discard    keep 
#    1100   11237 
# === sample_1MI3_so ===
# 
# discard    keep 
#    1275    8168 
# === sample_2MI1_so ===
# 
# discard    keep 
#    1538   10132 
# === sample_2MI3_so ===
# 
# discard    keep 
#    1031    8550 
# === sample_3C_MI2_so ===
# 
# discard    keep 
#    1134    8842 
# === sample_3MI1_so ===
# 
# discard    keep 
#     866    9743 
# === sample_3MI3_so ===
# 
# discard    keep 
#    1494    8224 
# === sample_4D_MI2_so ===
# 
# discard    keep 
#    1092    8559 
# === sample_4MI1_so ===
# 
# discard    keep 
#     858    6370 
# === sample_4MI3_so ===
# 
# discard    keep 
#    1289    8655 
# === sample_5MI1_so ===
# 
# discard    keep 
#    1062    8037 
# === sample_5MI3_so ===
# 
# discard    keep 
#    1245    8591 
# === sample_6MI1_so ===
# 
# discard    keep 
#    1447   10041 
# === sample_6MI3_so ===
# 
# discard    keep 
#    1202    7524 
# === sample_7MI1_so ===
# 
# discard    keep 
#     829    8835 
# === sample_7MI3_so ===
# 
# discard    keep 
#    1222    8104 
# === sample_8MI1_so ===
# 
# discard    keep 
#    1240    8584 
# === sample_8MI3_so ===
# 
# discard    keep 
#    1703   10719 





```


Now subset cells in each sample to only keep the cells that say keep in miQC.keep column of Seurat meta.data
(re-run 5/29/25)
```{r}

for (s in sample_names) {
  obj <- get(s)
  obj <- subset(obj, miQC.keep == "keep")
  assign(s, obj)
}

```


Calculate number of cells per sample now
(re-run 5/29/25)
```{r}
for (s in sample_names) {
  obj <- get(s)
  cat(s, ":", ncol(obj), "cells\n")
}


# sample_1MI1_so : 11237 cells
# sample_1MI3_so : 8168 cells
# sample_2MI1_so : 10132 cells
# sample_2MI3_so : 8550 cells
# sample_3C_MI2_so : 8842 cells
# sample_3MI1_so : 9743 cells
# sample_3MI3_so : 8224 cells
# sample_4D_MI2_so : 8559 cells
# sample_4MI1_so : 6370 cells
# sample_4MI3_so : 8655 cells
# sample_5MI1_so : 8037 cells
# sample_5MI3_so : 8591 cells
# sample_6MI1_so : 10041 cells
# sample_6MI3_so : 7524 cells
# sample_7MI1_so : 8835 cells
# sample_7MI3_so : 8104 cells
# sample_8MI1_so : 8584 cells
# sample_8MI3_so : 10719 cells


```


Add numbers of nuclei in each sample
(re-run 5/29/25)
```{r}
# sum(...): adds up all the cell counts across objects
# sapply(..., function): applies the function to each sample
# get(s): retrieves the actual Seurat object from the environment using its name
# ncol(get(s)): returns the number of columns in the object’s count matrix (i.e., the number of cells/nuclei)
total_nuclei <- sum(sapply(sample_names, function(s) ncol(get(s))))
total_nuclei
# 158,915 nuclei left


# double check
# > 11237+8168+10132+8550+8842+9743+8224+8559+6370+8655+8037+8591+10041+7524+8835+8104+8584+10719
# [1] 158915


# went from 180,542 nuclei to 158,915 nuclei
# lost 21627 nuclei so far


```


5/22/25 
Next steps:

Also subset the cells so that we are only keeping cells that have a UMI > 1000 (UMI count of 1000 or higher)
ALSO ADD TO THIS filter out cells that have < 5% mito reads and cells that have < 5% ribo reads
(re-run 5/29/25)
```{r}

# > sample_names
#  [1] "sample_1MI1_so"   "sample_1MI3_so"   "sample_2MI1_so"   "sample_2MI3_so"   "sample_3C_MI2_so" "sample_3MI1_so"   "sample_3MI3_so"   "sample_4D_MI2_so" "sample_4MI1_so"  
# [10] "sample_4MI3_so"   "sample_5MI1_so"   "sample_5MI3_so"   "sample_6MI1_so"   "sample_6MI3_so"   "sample_7MI1_so"   "sample_7MI3_so"   "sample_8MI1_so"   "sample_8MI3_so"


# Loop through each sample object, filter to keep cells that have UMI 1,000 or greater, less than 5% mito, and less than 5% ribi calculate percent.rb, and update the object
for (s in sample_names) {
  obj <- get(s)
  obj<-subset(obj, subset = nCount_RNA > 1000 & percent.mt < 5 & percent.rb < 5)
  assign(s, obj)
}


```


UMI count may be too stringent as already below 200k nuclei across all samples (goal was 200k high quality nuclei). Haven't run Seurat pre-processing steps or DoubletFinder yet



Re-calculate total nuclei before running standard Seurat steps and DoubletFinder
(re-run 5/29/25)
```{r}

# sum(...): adds up all the cell counts across objects
# sapply(..., function): applies the function to each sample
# get(s): retrieves the actual Seurat object from the environment using its name
# ncol(get(s)): returns the number of columns in the object’s count matrix (i.e., the number of cells/nuclei)
total_nuclei <- sum(sapply(sample_names, function(s) ncol(get(s))))
total_nuclei
# 146,635 nuclei left

# 158,915-146,635 =  12,280 removed

```


Next, run Seurat steps then run doublet finder. DF requires that all low quality cells are filtered out BEFORE running! Only use good quality cells
DF needs clustering information before it's run because it's going to simulate doublets in clusters, so PCA & UMAPs need to be created first
Refer to powerpoint on dropbox to see what previous UMAPs look like


Seurat pre-processing steps for each sample! 
Normalize data
FindVariableFeatures
ScaleData
RunPCA
ElbowPlot
FindNeighbors
FindClusters
RunUMAP


5/21/25

SKIP
Normalize data
```{r}

# Loop through each sample object to run Seurat preprocessing steps
# Using one function at a time
# for (s in sample_names) {
#   obj <- get(s)
#   obj<-NormalizeData(obj)
#   assign(s, obj)
# }

```


Pipe through each Seurat function with %<% (eqivalent to | on command line). Takes a little while because 18 samples!
(re-run 5/29/25)
```{r}

# Loop through each sample object to run Seurat preprocessing steps all in one loop
# default highly variable genes (HVG's) to calculate is 2000
# default resolution for FindClusters is 0.8
for (s in sample_names) {
  obj <- get(s)
  obj <- obj %>%
    NormalizeData() %>%
    FindVariableFeatures() %>%
    ScaleData() %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:30) %>%
    FindClusters() %>%
    RunUMAP(dims = 1:30)
  assign(s, obj)
}




```


5/29/25

Make elbow plots of samples to see if using PC 1-30 is ok

IMPORTANT: When just making plots with all 18 samples, DO NOT use assign(s, obj) in the loop, because it will overwrite the seurat objects (so's) with ggplot objects!!!!!!!!!!
```{r}
#######################################SKIP###################

# RAN THIS but it assigned the elbow plot to each sample so it overwrote my seurat objects...DO NOT RUN
# for (s in sample_names) {
#   obj <- get(s)
#   obj<-ElbowPlot(obj, ndims = 45)
#   assign(s, obj) <<<----- this line caused the issue. # this overwrites your Seurat object with a plot
#   
#   
# }
#######################################SKIP###################


# # output plots --- already overwritten :(
# for (s in sample_names) {
#   obj <- get(s)
#   print(ElbowPlot(obj, ndims = 45) + ggtitle(s))
# }

# now seurat objects replaced with ggplot objects.....should just print one at a time

# 30 pc's is good for now!
# sample_1MI1_so


# You’ll need to rerun the code that created the Seurat objects, or reload them from saved files, because obj <- ElbowPlot(...) has replaced them.


# see around line 440 on how to read in seurat objects....ran chunk and they are still ggplot elbow plot objects...:(
# see around line 467 for the fix


# re-run W/O assigning plot back to Seurat object....don't want it to get overwritten again
# IT WORKED!!!!!!!!
# use ggtitle to create title of sample so for each plot
for (s in sample_names) {
  obj <- get(s)
  print(ElbowPlot(obj, ndims = 45) + ggtitle(s))
}



```


IMPORTANT: When just making plots with all 18 samples, DO NOT use assign(s, obj) in the loop, because it will overwrite the seurat objects (so's) with ggplot objects!!!!!!!!!!



Forgot to make QC violin plots per sample!!!!
```{r}
# remove black dots with pt.size = 0
# will make cluster specific because already ran UMAP algorithm steps. If made violins before then it would be just one violin per sample QC metric
# make not smushed together? just expand plot window!
# also percent.rb didn't get plotted or is blocked by sample name
for (s in sample_names) {
  obj <- get(s)
  print(VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), ncol = 4, pt.size = 0) + ggtitle(s))
}

```




6/2/25
Run DoubletFinder on each sample using PCs 1-30
May need to run as 1 command per sample because each step requires different object name changes? no, will take too long
Refer to GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 1383



Ex highest pK without making graphs for each sample to see manually
See GB_MI5_macfas5_liftoff_multiomics_workflow_cellranger_arc_10_8_24.Rmd line 7498
# add pK_ prefix to DF objects 
# extract highest pK value from graphs above
pK_liftoff_1_MI5_V1_SO_filtered <- as.numeric(as.character(bcmvn_liftoff_1_MI5_V1_SO_filtered %>% filter(BCmetric == max(BCmetric)) %>% pull(pK)))
See GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd script where I ran doubletFinder lines 1142-1529

WODE LOOP WORKS :)
```{r}

# library(DoubletFinder)

for (s in sample_names) {
  obj <- get(s)
  
  ## pK Identification (no ground-truth) -------------------------
  sweep.res.list <- paramSweep(obj, PCs = 1:30, sct = FALSE)
  sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
  bcmvn <- find.pK(sweep.stats)
  
  ## Select the optimal pK value
  best.pK <- as.numeric(as.character(bcmvn[which.max(bcmvn$BCmetric), "pK"]))
  # best.pK <- as.numeric(as.character(obj %>% filter(BCmetric == max(BCmetric)) %>% pull(pK))) got error
    
  ## Homotypic doublet proportion estimate ------------------------
  annotations <- obj@meta.data$seurat_clusters
  homotypic.prop <- modelHomotypic(annotations)
  
  ## Number of expected doublets (adjust as needed) --------------
  nExp <- round(0.08 * ncol(obj))  # 8% doublet rate from 10x kit instructions
  nExp.adj <- round(nExp * (1 - homotypic.prop))
  
  ## Run DoubletFinder -------------------------------------------
  obj <- doubletFinder(obj, PCs = 1:30, pN = 0.25, pK = best.pK, nExp = nExp.adj, reuse.pANN = FALSE, sct = FALSE)
  
  # Store updated object
  assign(s, obj)
}



```




Check metadata of samples to make sure DF was run and you see pANN and DF.classifications columns (YES)
```{r}

# head() only prints 1st 6 lines of seurat object meta.data
for (s in sample_names) {
  cat("\nMetadata for", s, "\n")
  print(head(get(s)@meta.data))
}


```



Check DF classifications of each sample
```{r}


for (s in sample_names) {
  obj <- get(s)
  
  # Find the column name that contains the doublet classifications
  df_col <- grep("^DF.classifications", colnames(obj@meta.data), value = TRUE)
  
  # cat sample name, list sample in sample_names, and go to new line
  cat("\nSample:", s, "\n")
  
  # just in case: if sample doesn't have DF. classifications, print out No DoubletFinder...If sample has column, use table function to summarize
  if (length(df_col) == 0) {
    cat("  No DoubletFinder classifications found.\n")
  } else {
    print(table(obj@meta.data[[df_col]]))
  }
}


# save output w/o re-running chunk

# Sample: sample_1MI1_so 
# 
# Doublet Singlet 
#     821   10069 
# 
# Sample: sample_1MI3_so 
# 
# Doublet Singlet 
#     526    6535 
# 
# Sample: sample_2MI1_so 
# 
# Doublet Singlet 
#     710    8825 
# 
# Sample: sample_2MI3_so 
# 
# Doublet Singlet 
#     527    6539 
# 
# Sample: sample_3C_MI2_so 
# 
# Doublet Singlet 
#     600    7434 
# 
# Sample: sample_3MI1_so 
# 
# Doublet Singlet 
#     717    8854 
# 
# Sample: sample_3MI3_so 
# 
# Doublet Singlet 
#     506    6404 
# 
# Sample: sample_4D_MI2_so 
# 
# Doublet Singlet 
#     614    7617 
# 
# Sample: sample_4MI1_so 
# 
# Doublet Singlet 
#     467    5816 
# 
# Sample: sample_4MI3_so 
# 
# Doublet Singlet 
#     540    6681 
# 
# Sample: sample_5MI1_so 
# 
# Doublet Singlet 
#     588    7290 
# 
# Sample: sample_5MI3_so 
# 
# Doublet Singlet 
#     596    7409 
# 
# Sample: sample_6MI1_so 
# 
# Doublet Singlet 
#     686    8432 
# 
# Sample: sample_6MI3_so 
# 
# Doublet Singlet 
#     509    6372 
# 
# Sample: sample_7MI1_so 
# 
# Doublet Singlet 
#     636    7935 
# 
# Sample: sample_7MI3_so 
# 
# Doublet Singlet 
#     552    6838 
# 
# Sample: sample_8MI1_so 
# 
# Doublet Singlet 
#     594    7456 
# 
# Sample: sample_8MI3_so 
# 
# Doublet Singlet 
#     744    9198
```

Remove rows that say Doublet and only keep Singlets
```{r}

# for (s in sample_names) {
#   obj <- get(s)
#   
#   # Find the DF.classifications column
#   df_col <- grep("^DF.classifications", colnames(obj@meta.data), value = TRUE)
#   
#   if (length(df_col) == 1) {
#     # Keep only Singlet rows/cells
#     obj <- subset(obj, subset = obj@meta.data[[df_col]] == "Singlet")
#     
#     # Reassign the filtered object back to the original name
#     assign(s, obj)
#   } else {
#     cat("Skipping", s, "- No or multiple DF.classifications columns found.\n")
#   }
# }
# error - can't use subset twice

# won't work either because subset functions needs seurat objects - obj - as input
# for (s in sample_names) {
#   obj <- get(s)
# 
#   # Find the DF.classifications column
#   df_col <- grep("^DF.classifications", colnames(obj@meta.data), value = TRUE)
# 
#   if (length(df_col) == 1) {
#     # Keep only Singlet rows/cells
#     obj <- subset(obj@meta.data[[df_col]] == "Singlet")
# 
#     # Reassign the filtered object back to the original name
#     assign(s, obj)
#   } else {
#     cat("Skipping", s, "- No or multiple DF.classifications columns found.\n")
#   }
# }




# Run without explicitly using subset R function (WORKS!)
for (s in sample_names) {
  obj <- get(s)
  df_col <- grep("^DF.classifications", colnames(obj@meta.data), value = TRUE)
  if (length(df_col) == 1) {
    obj <- obj[, obj@meta.data[[df_col]] == "Singlet"]
    assign(s, obj)
  }
}




```


Check metadata again (last row should all be singlets!)
```{r}

# head() only prints 1st 6 lines of seurat object meta.data
for (s in sample_names) {
  cat("\nMetadata for", s, "\n")
  print(head(get(s)@meta.data))
}


```


Calculate again how many cells/nuclei are in each sample to get total
```{r}

# total nuclei before running doublet finder was # 146,635 nuclei left (line 783)
total_nuclei_afterdf <- sum(sapply(sample_names, function(s) ncol(get(s))))
total_nuclei_afterdf

# 135,704 nuclei left after running doublet finder
# > 146635-135704
# [1] 10931           10,931 nuclei removed

```



Now save fully QC'd objects in cluster BEFORE making some UMAP feature plots
```{r}


# # Output directory
outdir <- "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
# 
# # List of Seurat object variable names
# sample_names <- c(
#   "sample_1MI1_so", "sample_1MI3_so", "sample_2MI1_so", "sample_2MI3_so",
#   "sample_3C_MI2_so", "sample_3MI1_so", "sample_3MI3_so", "sample_4D_MI2_so",
#   "sample_4MI1_so", "sample_4MI3_so", "sample_5MI1_so", "sample_5MI3_so",
#   "sample_6MI1_so", "sample_6MI3_so", "sample_7MI1_so", "sample_7MI3_so",
#   "sample_8MI1_so", "sample_8MI3_so"
# )
# 
# # since alot of samples, save each Seurat object in for loop and attach date and short note on samples
# fully QC'd means filtered UMI, percent.ribo, percent.mt, ran miQC and DoubletFinder.
for (s in sample_names) {
  obj <- get(s)

  # Save with date in filename
  saveRDS(obj, file = file.path(outdir, paste0(s, "_fully_QCed_6_2_25.rds")))
}


# log into cluster and check ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files to see they are there


```


6/3/25
How to read in QC'd objects (reference line 467-482)
```{r}

# indir <- "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files"
# 
# # Only load the fully QCed files
# rds_files <- list.files(indir, pattern = "_fully_QCed_6_2_25\\.rds$", full.names = TRUE)
# 
# for (file in rds_files) {
#   obj <- readRDS(file)
#   obj_name <- gsub("_fully_QCed_6_2_25\\.rds$", "", basename(file))  # Strip suffix for object name
#   cat("Loaded:", obj_name, "\n")
#   print(dim(obj))
#   assign(obj_name, obj)
# }
# 
# # Create sample name list
# sample_names <- gsub("_fully_QCed_6_2_25\\.rds$", "", basename(rds_files))


```















View UMAPs + create gene violin/feature plots with the 18 samples. REMEMBER - WHEN JUST MAKING PLOTS, DO NOT USE assign(s, obj) in the loop! It will overwrite seurat objects with ggplot2 objects
Reference lines 912-915 as an example


View UMAPs with resolution 0.8 as default
```{r}

for (s in sample_names) {
  obj <- get(s)
  print(DimPlot(obj, reduction = "umap", label = TRUE, repel = TRUE) + ggtitle(s))
}


```



Make OSTN gene violin plots
```{r}

for (s in sample_names) {
  obj <- get(s)
  print(VlnPlot(obj, feature = "OSTN", pt.size = FALSE) + ggtitle(s))
}


# hmmm...seems lowly expressed except for a few samples (3MI1 cluster 16 and 4MI3 cluster 21?)
```

Make OSTN gene feature plots
```{r}

for (s in sample_names) {
  obj <- get(s)
  print(FeaturePlot(obj, feature = "OSTN") + ggtitle(s))
}

```


COMBINE ALL SEURAT OBJECTS

See how I can combine all seurat objects into one without having to do manually like 1st round
View Violeta's code Schaefer lab line 642 https://github.com/SchaferLabUMassChan/Duran-Laforet-V_2024/blob/main/MERFISH_analysis_allcells_Duran-Laforet_2024.R

Also view Seurat page https://satijalab.org/seurat/archive/v4.3/merge
By default, merge() will combine the Seurat objects based on the raw count matrices, erasing any previously normalized and scaled data matrices. If you want to merge the normalized data matrices as well as the raw count matrices, simply pass merge.data = TRUE. This should be done if the same normalization approach was applied to all objects.

```{r}

all_combined <- merge(
  sample_1MI1_so,
  y = c(
    sample_1MI3_so,
    sample_2MI1_so,
    sample_2MI3_so,
    sample_3C_MI2_so,
    sample_3MI1_so,
    sample_3MI3_so,
    sample_4D_MI2_so,
    sample_4MI1_so,
    sample_4MI3_so,
    sample_5MI1_so,
    sample_5MI3_so,
    sample_6MI1_so,
    sample_6MI3_so,
    sample_7MI1_so,
    sample_7MI3_so,
    sample_8MI1_so,
    sample_8MI3_so
  ),
  add.cell.ids = c(
    "sample_1MI1_so",
    "sample_1MI3_so",
    "sample_2MI1_so",
    "sample_2MI3_so",
    "sample_3C_MI2_so",
    "sample_3MI1_so",
    "sample_3MI3_so",
    "sample_4D_MI2_so",
    "sample_4MI1_so",
    "sample_4MI3_so",
    "sample_5MI1_so",
    "sample_5MI3_so",
    "sample_6MI1_so",
    "sample_6MI3_so",
    "sample_7MI1_so",
    "sample_7MI3_so",
    "sample_8MI1_so",
    "sample_8MI3_so"
  ),
  merge.data = TRUE
)

```

It worked!!!!!! object has 135,704 nuclei!!!! (exactly number of nuclei left after QC)
```{r}

all_combined


# An object of class Seurat 
# 23477 features across 135704 samples within 1 assay 
# Active assay: RNA (23477 features, 2000 variable features)
#  54 layers present: counts.1, counts.2, counts.3, counts.4, counts.5, counts.6, counts.7, counts.8, counts.9, counts.10, counts.11, counts.12, counts.13, counts.14, counts.15, counts.16, counts.17, counts.18, data.1, scale.data.1, data.2, scale.data.2, data.3, scale.data.3, data.4, scale.data.4, data.5, scale.data.5, data.6, scale.data.6, data.7, scale.data.7, data.8, scale.data.8, data.9, scale.data.9, data.10, scale.data.10, data.11, scale.data.11, data.12, scale.data.12, data.13, scale.data.13, data.14, scale.data.14, data.15, scale.data.15, data.16, scale.data.16, data.17, scale.data.17, data.18, scale.data.18




```

Check rows/cell id's
# notice the cell names now have an added identifier
```{r}

head(colnames(all_combined))
tail(colnames(all_combined))


# [1] "sample_1MI1_so_AAACCCAAGAGACAAG-1" "sample_1MI1_so_AAACCCAAGTCCCGAC-1" "sample_1MI1_so_AAACCCAAGTGAGGCT-1" "sample_1MI1_so_AAACCCACAGAAGCTG-1"
# [5] "sample_1MI1_so_AAACCCACATCTATCT-1" "sample_1MI1_so_AAACCCACATGGAGAC-1"
# [1] "sample_8MI3_so_TTTGTTGAGGTTAGTA-1" "sample_8MI3_so_TTTGTTGCAACGGTAG-1" "sample_8MI3_so_TTTGTTGCAAGAAACT-1" "sample_8MI3_so_TTTGTTGGTAACATGA-1"
# [5] "sample_8MI3_so_TTTGTTGGTTACCCTC-1" "sample_8MI3_so_TTTGTTGTCTATACTC-1"



```

Save combined object as RDS!
```{r}

# saveRDS(all_combined, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_before_preprocess_6_3_25.rds")

```



Reference lines 3261 down in GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd
NEXT re-run the standard seurat steps again with combined object (except normalize data)!!!
```{r}

# pipe through seurat functions like previously!
# learned from scRNAseq class: If re-running analysis, no need to run NormalizeData() again

# Why you can skip NormalizeData() after merging:
# You already ran NormalizeData() on each sample individually, which means:
# 
# The raw counts for each sample have already been log-normalized (or normalized per whatever method you chose).
# 
# After merging, the RNA assay in the merged object retains the normalized data in the "data" slot and raw counts in the "counts" slot.
# 
# Running NormalizeData() again on the merged object would:
# 
# Double-normalize your data — which is not mathematically valid.
# 
# Overwrite your existing normalized values.
# 
# Potentially introduce unwanted artifacts, especially if your samples had different library sizes or sequencing depths.



```


View seurat object metadata
```{r}

View(all_combined@meta.data)

```


Run on combined object
```{r}

all_combined <- all_combined %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)

```



Plot UMAP!
```{r}

# resolution 0.8
DimPlot(all_combined, reduction = "umap", label = TRUE)
# 43 clusters
# save as png: GB_combined_rhemac10_snRNAseq_UMAP_PC1-30_res0.8_6_5_25.png

```

Added 6/12/25
Calculate number of cells per cluster! Did in GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd line 5198
```{r}

# use table function for cell cluster id's (Idents)
table(Idents(all_combined))


#    0     1     2     3     4     5     6     7     8     9    10    11    12    13    14    15    16    17    18    19    20    21    22    23    24    25    26    27 
# 13472 11764 11269 11183 10175  8189  6772  5765  5443  4993  4894  4873  4192  3201  2850  2704  2587  2576  2420  2039  1851  1822  1412  1212  1135  1061   896   742 
#    28    29    30    31    32    33    34    35    36    37    38    39    40    41    42 
#   723   607   521   519   276   257   244   193   187   170   160   130    94    88    43 

```


6/6/26
Make QC violin plt (not removing any nuclei, just to see)
jk hard to see because there's 43 clusters
```{r}

# hard to see because number of clusters, make one at a time
# VlnPlot(all_combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), ncol = 4, pt.size = 0)


VlnPlot(all_combined, features = "nFeature_RNA", ncol = 4, pt.size = 0)


VlnPlot(all_combined, features = "nCount_RNA", ncol = 4, pt.size = 0)


VlnPlot(all_combined, features = "percent.mt" , ncol = 4, pt.size = 0)


VlnPlot(all_combined, features = "percent.rb", ncol = 4, pt.size = 0)


```







Make gene violin plots (only some...hard to see) and feature plots with genes from GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd
Since used ensembl GTF NOT ALL genes have gene symbol. Seems that most genes below are in GTF!
Reference lines 1620-1640, 3354-3448
```{r}

# reminder that using PC1-30 and resolution 0.8 (higher number PC's and resolution than 1st time)


FeaturePlot(all_combined, features = c("SLC17A7", "RORB", "FEZF2", "HPCAL1", "THEMIS", "NXPH4", "OSTN","NPY", "GAD1", "LINC00473"))

FeaturePlot(all_combined, features = c("LHX6", "NR2F2", "SST", "PVALB", "VIP", "LAMP5", "MOG", "PDGFRA", "AQP4", "EGR1"))

FeaturePlot(all_combined, features = c("FLT1", "C1QB", "FOS", "EGR1", "NPAS4", "LINC00507", "ZNF331", "PER1", "IGF1", "BDNF"))

# Coritcal Excitatory Neurons/Layers
FeaturePlot(all_combined, features = c("SATB2", "POU3F3","POU3F2", "CUX2", "TBR1", "ETV1", "BCL11B", "CUX1", "FOXP1", "OTX1", "RELN"))


# Cortical GABAergic Neurons
FeaturePlot(all_combined, features = c("GAD2", "GAD1","SLC32A1", "CALB1", "CALB2", "SST", "VIP", "PVALB", "GABRB1", "GABRB2", "GABRB3"))


# Astrocytes
FeaturePlot(all_combined, features = c("GFAP", "S100B","AQP4"))

# Oligodendrocytes
FeaturePlot(all_combined, features = c("OLIG2", "MBP"))

# Microglia
FeaturePlot(all_combined, features = c("ITGAM", "PTPRC"))


```


Split up UMAP's so can see better
Keep at 4 genes per page for lines 1485-1489 above
```{r}

# Group 1
FeaturePlot(all_combined, features = c("SLC17A7", "RORB", "FEZF2", "HPCAL1"))

# Group 2
FeaturePlot(all_combined, features = c("THEMIS", "NXPH4", "OSTN", "NPY"))

# Group 3
FeaturePlot(all_combined, features = c("GAD1", "LINC00473", "LHX6", "NR2F2"))
# LINC00473 isn't symbol in ensembl GTF. Would need to find mmul_10 ensembl id

# Group 4
FeaturePlot(all_combined, features = c("SST", "PVALB", "VIP", "LAMP5"))

# Group 5
FeaturePlot(all_combined, features = c("MOG", "PDGFRA", "AQP4", "EGR1"))

# Group 6
FeaturePlot(all_combined, features = c("FLT1", "C1QB", "FOS", "EGR1"))

# Group 7
FeaturePlot(all_combined, features = c("NPAS4", "LINC00507", "ZNF331", "PER1"))
# LINC00507 isn't symbol in ensembl GTF. Would need to find mmul_10 ensembl id

# Group 8
FeaturePlot(all_combined, features = c("IGF1", "BDNF"))



# save plots as png's and put in lab dropbox BoultingLab_shared / Experiments / snRNAseq Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF / all_plots_tables_created.pptx


```

Cortical Excitatory Neurons
```{r}

FeaturePlot(all_combined, features = c("SATB2", "POU3F3", "POU3F2", "CUX2"))
FeaturePlot(all_combined, features = c("TBR1", "ETV1", "BCL11B", "CUX1"))
FeaturePlot(all_combined, features = c("FOXP1", "OTX1", "RELN"))


```


Cortical GABAergic Neurons
```{r}


FeaturePlot(all_combined, features = c("GAD2", "GAD1", "SLC32A1", "CALB1"))
FeaturePlot(all_combined, features = c("CALB2", "SST", "VIP", "PVALB"))
FeaturePlot(all_combined, features = c("GABRB1", "GABRB2", "GABRB3"))

# PVALB & VIP expressed in different clusters!

```



```{r}

# got these genes from Wei et al 2022 paper figure 1
FeaturePlot(all_combined, features = c("C1QB", "FLT1", "PDGFRA"))

# other markers
FeaturePlot(all_combined, features = c("FOXG1", "FOXP2"))


```


Make gene violin plot of OSTN
```{r}

VlnPlot(all_combined, features = "OSTN", pt.size = FALSE)
# highly expressed in cluster 21
# save as png gb_mmul10_snRNAseq_geneviolin_OSTN.png

```

Save seurat object since seurat preprocess has been completed for now
```{r}

saveRDS(all_combined, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_6_6_25.rds")


```



6/12/25
More feature plots!
```{r}
#  NPR3 gene may be the receptor for OSTN?
FeaturePlot(all_combined, features = c("MEF2C", "NPR3", "ARHGEF2"))


```


6/16/25
Review combined seurat object of 18 libraries
```{r}

all_combined

```






Run Neuroestimator on full snRNAseq dataset! Run on cluster and not in R. Reference GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 4746-4773 (extract counts matrix from R object), lines 4890-5068 (run NE in cluster)

FIRST, extract counts matrix from all combined R object. Save it as an rds object
Make folder in cluster ~/macaque_snRNAseq called neuroestimator to put results in
```{r}
# using Seurat v5, which introduces multi-layer assays. In this version, GetAssayData() no longer works directly on an assay with multiple layers unless you specify the layer  # all_combined_count_matrix <- GetAssayData(object = all_combined, assay = "RNA", layer = "counts")
                                                  

# all_combined_count_matrix <- all_combined[["RNA"]]$counts1, counts2, counts3 ... counts 18 (won't work)

# only 1st sample counts matrix is extracted....want to extract all 18 counts matrices...all_combined object contains 18 merged seurat objects
# all_combined_count_matrix <- LayerData(object = all_combined, assay = "RNA", layer = "counts")

# remove all_combined_count_matrix object

```


List object layers - want to extract raw counts layers - 18 of them b/c 18 samples
```{r}
# List of object layers
Layers(all_combined)

```


Manually combine count layers using layers function in Seurat https://satijalab.github.io/seurat-object/reference/Layers.html
Count matrix should have 23,447 rows/genes/features and 135,704 samples/columns/nuclei
```{r}

# List all layers that start with "counts"
all_counts_layers <- Layers(all_combined[["RNA"]])
all_counts_layers <- counts_layers[grepl("^counts", all_counts_layers)]

# # view all_count_layers
#  [1] "counts.1"  "counts.2"  "counts.3"  "counts.4"  "counts.5"  "counts.6"  "counts.7"  "counts.8"  "counts.9"  "counts.10" "counts.11" "counts.12" "counts.13" "counts.14"
# [15] "counts.15" "counts.16" "counts.17" "counts.18"



# Extract each layer and combine
all_count_matrices <- lapply(all_counts_layers, function(layer) {
  LayerData(all_combined, assay = "RNA", layer = layer)
})


# View all_count_matrices (prints out all 18 count matrices....each matrices has different number of rows/genes and different number of columns....)
# all_count_matrices



# Combine into one gene x cell matrix
all_combined_counts <- do.call(cbind, all_count_matrices)

# Error in cbind.Matrix(x, y, deparse.level = 0L) : 
#   number of rows of matrices must match


```


In order to combine all count matrices, all rows (genes) must total to 23,447. Check below
```{r}
# Get all gene names from all 18 count matrices
all_genes <- unique(unlist(lapply(all_count_matrices, rownames)))

# check how many total genes as sanity check
length(all_genes) 


```


```{r}

# Get reference gene order from the first matrix
gene_order <- rownames(all_count_matrices[[1]])

# Reorder each matrix to this order
aligned_matrices <- lapply(all_count_matrices, function(mat) {
  mat[gene_order, , drop = FALSE]  # Ensure same order
})


all_combined_counts <- do.call(cbind, aligned_matrices)


```


Opened issue on Seurat github & biostars


Try this (didn't work :())
```{r}

test <- LayerData(object = all_combined, assay = "RNA", layer = c("counts.1", "counts.2"))
# Warning: multiple layers are identified by counts.1 counts.2
#  only the first layer is used

```

Try this
```{r}

test2 <- LayerData(object = all_combined, assay = "RNA", layer = "counts", search = NA)



```



See this biostars issue https://www.biostars.org/p/9586642/
Data are stored as individual layers in Seurat V5 object that's why you see 6 layers even though you have merged them. So, if you would like to a single combined layer, you can simply Join layer by following function JoinLayers()


See seurat docs here https://satijalab.org/seurat/articles/seurat5_integration
Seurat v5 assays store data in layers. These layers can store raw, un-normalized counts (layer='counts'), normalized data (layer='data'), or z-scored/variance-stabilized data (layer='scale.data').

Once integrative analysis is complete, you can rejoin the layers - which collapses the individual datasets together and recreates the original counts and data layers. You will need to do this before performing any differential expression analysis. However, you can always resplit the layers in case you would like to reperform integrative analysis





JoinLayers() https://satijalab.github.io/seurat-object/reference/SplitLayers.html

Seurat layers are individual samples

Join layers (samples) in all_combined. SAVE TO NEW SEURAT object so can keep original file
```{r}

all_combined_join_layers <- JoinLayers(all_combined)


```



View new seurat object - combined the 18 counts layers into one counts and data layers into one data. The scale.data is stil split into the 18 samples
raw, un-normalized counts (layer='counts'), normalized data (layer='data'), or z-scored/variance-stabilized data (layer='scale.data').
```{r}

all_combined_join_layers

```


Join scale.data layers? No just splits everything up again
```{r}

# all_combined_join_layers <- JoinLayers(all_combined, layers = "scale.data")

```



Ok now with new seurat object, extract counts matrix - DID IT!!!!!!!! Update issues I opened lol and close them
```{r}

all_combined_count_matrix_2 <- LayerData(object = all_combined_join_layers, assay = "RNA", layer = "counts")


```



View combined counts matrix
```{r}

all_combined_count_matrix_2


```



Now save combined counts matrices as an rds object in ~/macaque_snRNAseq/neuroestimator
Remember that this is a sparse matrices. Need to convert it to a normal matrix using as.matrix function in R on cluster in order for neuroestimator to work
```{r}

# accidentally saved as same name as original combined object just in different folder. Resave below
# saveRDS(all_combined_count_matrix_2, file = "~/macaque_snRNAseq/neuroestimator/all_combined_counts_matrix_6_16_25.rds")

```


NEXT, RUN NE on cluster!!!

Get working directory (pwd)
/home/gb1-umw/macaque_snRNAseq/neuroestimator

Start interactive session
bsub -Is -q interactive -R "rusage[mem=64G]" /bin/bash

Activate conda env with neuroestimator installed
conda activate neuroestimator

Open R in command line
R

Get working directory in R
getwd()

Read in sparse count matrix
all_combined_count_matrix <- readRDS("all_combined_counts_matrix_6_16_25.rds")


Convert sparse matrix to normal matrix (dots become zero's) [UPDATE - DO IN RMD SCRIPT ITSELF, NOT ON CLUSTER CLI]
all_combined_count_matrix <- as.matrix(all_combined_count_matrix)


Got error
Error in asMethod(object) : 
  Cholmod error 'problem too large' at file ../Core/cholmod_dense.c, line 102
  
Request more memory in interactive job. Also specify multiple nodes to request memory per node instead of just one node.

When exiting R, don't save workspace (takes too long)



Load neuroestimator
library(neuroestimator)


Run neuroestimator



6/18/25

Started new RStudio server session Start new job (long, 720h, 32 cores, 9200MB > started quickly!!! Job id 385875. Total memory 32 x 9.2 = 294.4 GB




Read in all_combined_count_matrix_2
```{r}

# all_combined_count_matrix_2 <- readRDS("~/macaque_snRNAseq/neuroestimator/all_combined_counts_matrix_6_16_25.rds")

```

View sparse matrix (neuroestimator doesn't work on sparse matrices)
```{r}
all_combined_count_matrix_2
```


Resave sparse matrix below
```{r}

saveRDS(all_combined_count_matrix_2, file = "~/macaque_snRNAseq/neuroestimator/all_combined_counts_matrix_2_06_18_25.rds")

```


Try converting sparse matrix into dense or full matrix here instead of cluster
Worked!!!!
```{r}
# save to new object name
all_combined_count_matrix_2_reg_metrix <- as.matrix(all_combined_count_matrix_2)

# just print 1st 6 lines, don't print entire matrix
head(all_combined_count_matrix_2_reg_metrix)

```


Save regular non-sparse matrix in cluster /home/gb1-umw/macaque_snRNAseq/neuroestimator
Sparse matrix is 3.4GB
Regular non sparse matrix is 25.5GB 
```{r}

saveRDS(all_combined_count_matrix_2_reg_metrix, file = "~/macaque_snRNAseq/neuroestimator/all_combined_count_matrix_2_reg_matrix_06_18_25.rds")

# should be able to just read this object into R on cluster and run neuroestimator on cluster!

```


Check dimensions of regular non sparse matrix
```{r}

dim(all_combined_count_matrix_2_reg_metrix)

```


Ran ne on cluster in R script and used shell script to run non-interactively!!!! finally FINISHED RUNNING!!!!!!


6/23/25
remove regular matrix from r session in console
rm(all_combined_count_matrix_2_reg_metrix)

Read in neuroestimator results csv ~/macaque_snRNAseq/neuroestimator/neuroestimator_results_rhemac10_snRNAseq_all_samples_6_18_25.csv
```{r}

neuroestimator_results <- read.csv(file = "~/macaque_snRNAseq/neuroestimator/neuroestimator_results_rhemac10_snRNAseq_all_samples_6_18_25.csv")

# see first 6 lines of results
head(neuroestimator_results)

# confirm that object type is a data.frame
class(neuroestimator_results)

# confirm that there are 135,704 rows for the 135,704 cells/nuclei
nrow(neuroestimator_results)

```



Read in all samples Seurat object (reference line 1624)
all_combined_count_matrix_2 is just the counts matrix, not a Seurat object
```{r}

all_combined <- readRDS(file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_6_6_25.rds")


```


```{r}

all_combined

```


Before adding csv results to seurat object, make historgam of neuroestimator scores using ggplot!
```{r}
# x-axis is predicted_activity % from neuroestimator.
# y-axis is number of cells/nuclei that fall into each bin (NOT the total number of nuclei on y-axis)
# default bin is 30 (data divided up into 30 equally spaced bins)
# aes() function specifies the aesthetic mappings, which maps predicted_activity variable to the x-axis
ggplot(data = neuroestimator_results) + 
  geom_histogram(mapping = aes(x = predicted_activity))

```


Edit histogram
https://www.datacamp.com/tutorial/make-histogram-ggplot2
https://www.atlassian.com/data/charts/histogram-complete-guide

```{r}

# customize the color of the outlines of each bar using the color attribute, and we can change the fill of the bars using the fill attribute of geom_histogram(). We will fill the bars with blue and change the outline color to white.

ggplot(data = neuroestimator_results) + 
  geom_histogram(mapping = aes(x = predicted_activity), color = "white", fill = "blue") +
  labs(x = 'Predicted Activity (%)', y = 'Nuclei', title = 'Rhemac10 snRNAseq Neuroestimator Results')

# save as a pnd & add to dropbox BoultingLab_shared / Experiments / snRNAseq Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF

# Rhemac10_snRNAseq_Neuroestimator_Results_6_23_25.png

# distribution skewed to the left

```




What's the highest and lowest activity score in neuroestimator results? What is the average activity score?
```{r}
# smallest score
min(neuroestimator_results$predicted_activity)
# [1] 0.2691268


# largest score
max(neuroestimator_results$predicted_activity)
# [1] 0.9075024


# average score
mean(neuroestimator_results$predicted_activity)
# [1] 0.4811709
```


What are the score ranges in the top/bottom 10% of all cells?
```{r}

# Extract predicted_activity column
scores <- neuroestimator_results$predicted_activity

# Bottom 10% threshold (i.e., 10th percentile) using quantile function
bottom_10_threshold <- quantile(scores, 0.10)

# Top 10% threshold (i.e., 90th percentile) using quantile function
top_10_threshold <- quantile(scores, 0.90)

# Get actual score ranges
# calculate range of scores that are less than or equal to the 10th percentile
bottom_10_range <- range(scores[scores <= bottom_10_threshold])
# # calculate range of scores that are greater than or equal to the 90th percentile
top_10_range <- range(scores[scores >= top_10_threshold])

# Output
cat("Bottom 10% score range:\n", bottom_10_range, "\n")
cat("Top 10% score range:\n", top_10_range, "\n")

# Bottom 10% score range:
#  0.2691268 0.3841956 
# Top 10% score range:
#  0.5967884 0.9075024 

```




Add neuroestimator_results to all_combined seurat object to make a UMAP
Reference GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 4934-5069
```{r}

# add metadata column to seurat object. make sure tro save RDS object as all_combined_seurat_preprocess_neuroestimator_6_23_25.rds
all_combined[["predicted_activity"]] <- neuroestimator_results[["predicted_activity"]]

# view seurat object meta.data now. There should be a predicted activity column
View(all_combined@meta.data)

```



Now make a feature UMAP plot of the activity scores
```{r}
# This is similar to how I colored the UMAP by nCount_RNA with the red dots. See lines 596-621 in GB_Mmul_10_3'_snRNAseq_workflow_scrublet.Rmd script

FeaturePlot(all_combined, features = c("predicted_activity"))

# save this UMAP on local computer and name it rhemac10_snRNAseq_all_combined_UMAP_neuroestimator_scores_6_23_25.png

```


What's the distribution of the activity scores in each cluster?
Would be helpful to see the distribution of the activity score in a violin plot and/or a bell curve plot!
See lines 3707-3715 for an example
```{r}

VlnPlot(all_combined, features = c("predicted_activity"), pt.size = FALSE) & theme(plot.title = element_text(size=10), axis.text=element_text(size=8))

# save to computer and name it rhemac10_snRNAseq_all_combined_violin_plot_neuroestimator_scores_6_23_25.png

# from looking at plot, clusters 1, 11, 14, 16, 28, 36, and 41 have the highest neuroestimator scores. Clsuter 36 seem to have the highest scores
```

Make bell curve/ridge plot also!
```{r}

RidgePlot(all_combined, features = c("predicted_activity"))

# looks so cool!!!! Export this as a png file and name it rhemac10_snRNAseq_all_combined_ridge_plot_neuroestimator_scores_6_23_25.png

```


Subset seurat object for cluster 2 & 21 (where OSTN expression is the highest from viewing UMAP)
Reference GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 3979-3991 on subsetting seurat objects based on cluster number
```{r}

# subsetting clusters as computational zoom in
all_combined_clusters_2_21 <- subset(all_combined, idents = c(2, 21))

# view number of cells in cluster 2 and 21 (should be 11269 + 1822 = 13091)
all_combined_clusters_2_21

```


Make histogram of ne scres again but just for clusters 2 and 21 to see distribution! (neuroestimator_results is just csv with no cluster information)
Has similar distro to entire dataset
```{r}


ggplot(data = all_combined_clusters_2_21@meta.data) + 
  geom_histogram(mapping = aes(x = predicted_activity), color = "white", fill = "blue") +
  labs(x = 'Predicted Activity (%)', y = 'Nuclei', title = 'Rhemac10 snRNAseq cluster 2 & 21 Neuroestimator Results')


# save as png rhemac10_snRNAseq_cluster_2_&_21_Neuroestimator_Results_hist.png

```




Also make violin plots of ne scores cluster 2 and 21
```{r}
VlnPlot(all_combined_clusters_2_21, features = c("predicted_activity"), pt.size = FALSE) & theme(plot.title = element_text(size=10), axis.text=element_text(size=8))

# save as png rhemac10_snRNAseq_cluster_2_&_21_Neuroestimator_Results_violins.png

```



Calc min, max, and mean score for just clusters 2 & 21
```{r}

# min score
min(all_combined_clusters_2_21@meta.data$predicted_activity)
# [1] 0.2981832

# max score
max(all_combined_clusters_2_21@meta.data$predicted_activity)
# [1] 0.8023094

# mean score
mean(all_combined_clusters_2_21@meta.data$predicted_activity)
# [1] 0.470414

```



Make UMAP of just clusters 2 and 21. Then make feature plot with OSTN gene
```{r}

DimPlot(all_combined_clusters_2_21, reduction = "umap", label = TRUE)

# save as png rhemac10_snRNAseq_cluster_2_&_21_Neuroestimator_Results_UMAP.png

```


```{r}

FeaturePlot(all_combined_clusters_2_21, features = 'OSTN')

# save as png rhemac10_snRNAseq_cluster_2_&_21_Neuroestimator_Results_OSTN_feature plot.png

```


6/24/25

Save all_combined object with neuroestimator predicted_activity metadata column added
```{r}
# save snRNAseq r object with neuroestimator metadata added
saveRDS(all_combined, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_neuroestimator_6_24_25.rds")


# read in snRNAseq r object with neuroestimator metadata added
# all_combined <- readRDS(file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_neuroestimator_6_24_25.rds")

```




- subset out just cluster 21 (since it has highest OSTN expression).

```{r}

# subsetting clusters as computational zoom in
all_combined_cluster_21 <- subset(all_combined, idents = 21)

# view number of cells (1822)
all_combined_cluster_21

```

- extract matrix from cluster 21 & send to Ag to run scikit learn python package to use k means clustering to check how it clusters

```{r}

# only 151 columns...
# all_combined_cluster21_count_matrix <- LayerData(object = all_combined_cluster_21, assay = "RNA", layer = "counts")
# all_combined_cluster21_count_matrix


# have to join layers 1st before extracting count matrix
# View new seurat object - combined the 18 counts layers into one counts and data layers into one data. The scale.data is stil split into the 18 samples
# raw, un-normalized counts (layer='counts'), normalized data (layer='data'), or z-scored/variance-stabilized data (layer='scale.data').
all_combined_cluster_21_join <- JoinLayers(all_combined_cluster_21)


# extract counts matrix from joined object (should be 1822 rows). Yes!
 all_combined_cluster21_count_matrix <- LayerData(object = all_combined_cluster_21_join, assay = "RNA", layer = "counts")
 all_combined_cluster21_count_matrix
 
```


Save as rds object. Use FZ to copy to UMMS computer & slack to Ag

```{r}

saveRDS(all_combined_cluster21_count_matrix, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_cluster21_count_matrix_6_24_25.rds")

```



- subcluster cluster 21 with seurat object all_combined_cluster_21 and make feature plot to see where OSTN is expressed (without NormalizeData step - entire dataset already normalized)
Refer to GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 3972-4217
```{r}

all_combined_cluster_21 <- all_combined_cluster_21 %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)



```


Plot UMAP
```{r}
# resolution 0.8
DimPlot(all_combined_cluster_21, reduction = "umap", label = TRUE)

# save as all_combined_cluster_21_umap.png
```


Make feature plot of OSTN
```{r}

FeaturePlot(all_combined_cluster_21, feature = "OSTN")
# save as all_combined_cluster_21_umap_OSTN.png

# top could be active cells and bottom can be inactive?!

```


View object metadata to make sure predicted_activity column is there (yes)
```{r}

View(all_combined_cluster_21@meta.data)

```


How can I see the scores of cells that highly express OSTN?

FetchData function Seurat https://satijalab.github.io/seurat-object/reference/FetchData.html
Retrieves data (feature expression, PCA scores, metrics, etc.) for a set of cells in a Seurat object
vars = List of variables to fetch, use keyword “ident” to pull identity classes
```{r}

# 1. Extract OSTN expression values
ostn_expr <- FetchData(all_combined_cluster_21, vars = "OSTN")

# 2. Combine with metadata
meta_with_expr <- cbind(all_combined_cluster_21@meta.data$predicted_activity, OSTN = ostn_expr$OSTN)

# 3. Filter cells with high OSTN expression (you can define a threshold)
# For example, top 10% expressing OSTN (90%):
threshold <- quantile(meta_with_expr$OSTN, 0.9)
high_ostn_cells <- meta_with_expr[meta_with_expr$OSTN >= threshold, ]

# 4. View neuroestimator scores (predicted_activity) of those cells
high_scores <- high_ostn_cells$predicted_activity



```



Try again (worked!)
```{r}
# Fetch both OSTN expression from matrix and predicted_activity
df <- FetchData(all_combined_cluster_21, vars = c("OSTN", "predicted_activity"))

# Set threshold for high OSTN expression — top 10% or 90% (can change to 95%, etc)
threshold <- quantile(df$OSTN, 0.9)

# Subset cells with high OSTN expression
# store into new object
# subset data frame (df) 
high_ostn_df <- df[df$OSTN >= threshold, ]

# View or summarize predicted_activity in these cells
summary(high_ostn_df$predicted_activity)


 # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 # 0.3640  0.4449  0.4814  0.5040  0.5711  0.7185

# median activity score for top 90% cells expressing OSTN is ~48%
# highest score is ~72%
# lowest score is ~36%

```



Check what's stored in variables below
```{r}

# both have 3 columns. 1st column has cell id's, 2nd column has normalized gene expression values for OSTN, and 3rd column has predicted_activity scores
head(df)
head(high_ostn_df)



```


Create histogram to see distribution!
```{r}

hist(high_ostn_df$predicted_activity,
     main = "Neuroestimator scores top 90% cells expressing OSTN",
     xlab = "Predicted Activity (%)",
     col = "skyblue")

# save as all_combined_cluster_21_histogram.png

```


How many cells express OSTN in cluster 21?

# FetchData can access anything from expression matrices, cell embeddings, or metadata Use the previously listed
# commands to access entire matrices Use FetchData to access individual/small groups of variables
FetchData(object = pbmc, vars = c("PC_1", "nFeature_RNA", "MS4A1"), layer = "counts")

```{r}

# Fetch OSTN expression values for all cells in the object
ostn_expr <- FetchData(all_combined_cluster_21, vars = "OSTN", layer = "counts")

# Count how many cells have non-zero OSTN expression
num_expressing <- sum(ostn_expr$OSTN > 0)

# Total number of cells in the object
total_cells <- ncol(all_combined_cluster_21)

# Percent expressing
percent_expressing <- (num_expressing / total_cells) * 100

# Output results
cat("Number of OSTN+ cells:", num_expressing, "\n")
cat("Total cells:", total_cells, "\n")
cat("Percent expressing OSTN:", round(percent_expressing, 2), "%\n")


# Number of OSTN+ cells: 702 
# Total cells: 1822 
# Percent expressing OSTN: 38.53 %

```



ADDED 6/30/25
Make neuroestimator histogram of just cluster 21 & calculate summary stats with summary() function & add to dropbox (do same for cluster 2)
```{r}

# calculate summary stats of neuroestimator scores
summary(all_combined_cluster_21@meta.data$predicted_activity)



ggplot(data = all_combined_cluster_21@meta.data) + 
  geom_histogram(mapping = aes(x = predicted_activity), color = "white", fill = "blue") +
  labs(x = 'Predicted Activity (%)', y = 'Nuclei', title = 'Rhemac10 snRNAseq cluster 21 Neuroestimator Results')


# save as rhemac10_snRNAseq_cluster_21_neuroestimator_results_hist.png



```






6/25/25
Subcluster cluster 2 also just to see what it looks like
```{r}

# subsetting clusters as computational zoom in
all_combined_cluster_2 <- subset(all_combined, idents = 2)

# view number of cells - 11269
all_combined_cluster_2

```


Subcluster
```{r}

all_combined_cluster_2 <- all_combined_cluster_2 %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)

```



Plot UMAP & feature plot
```{r}


# resolution 0.8
DimPlot(all_combined_cluster_2, reduction = "umap", label = TRUE)
# save as all_combined_cluster_2_umap.png


FeaturePlot(all_combined_cluster_2, feature = "OSTN")
# save as all_combined_cluster_2_umap_OSTN.png

# kind of has cool fan pattern line subcluster 4 in the previous round of snRNAseq analysis!
# left side could be active and right side could be inactive?


```


Make histogram
```{r}

# Fetch both OSTN expression from matrix and predicted_activity
df <- FetchData(all_combined_cluster_2, vars = c("OSTN", "predicted_activity"))

# Set threshold for high OSTN expression — top 10% or 90% (can change to 95%, etc)
threshold_2 <- quantile(df$OSTN, 0.9)

# Subset cells with high OSTN expression
# store into new object
# subset data frame (df) 
high_ostn_df_2 <- df[df$OSTN >= threshold, ]

# View or summarize predicted_activity in these cells
summary(high_ostn_df_2$predicted_activity)

 #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 # 0.3347  0.4202  0.4655  0.4769  0.5255  0.6629 


# median activity score for top 90% cells expressing OSTN is ~47%
# highest score is ~66%
# lowest score is ~33%

# changed title slightly to Neuroestimator scores top 90% cells expressing OSTN Subcluster 2 but title is too long
hist(high_ostn_df$predicted_activity,
     main = "Neuroestimator scores top 90% cells expressing OSTN",
     xlab = "Predicted Activity (%)",
     col = "skyblue")

# save as all_combined_cluster_2_histogram.png
# can kind of see a bimodal distribution too

```


How many cells express OSTN in cluster 2?
```{r}

# Fetch OSTN expression values for all cells in the object
ostn_expr <- FetchData(all_combined_cluster_2, vars = "OSTN", layer = "counts")

# Count how many cells have non-zero OSTN expression
num_expressing <- sum(ostn_expr$OSTN > 0)

# Total number of cells in the object
total_cells <- ncol(all_combined_cluster_2)

# Percent expressing
percent_expressing <- (num_expressing / total_cells) * 100

# Output results
cat("Number of OSTN+ cells:", num_expressing, "\n")
cat("Total cells:", total_cells, "\n")
cat("Percent expressing OSTN:", round(percent_expressing, 2), "%\n")

# Number of OSTN+ cells: 912 
# Total cells: 11269 
# Percent expressing OSTN: 8.09 %

```


ADDED 6/30/25
Make neuroestimator histogram of just cluster 2 & calculate summary stats with summary() function & add to dropbox (Did same for cluster 21 above)


```{r}

# calculate summary stats of neuroestimator scores
summary(all_combined_cluster_2@meta.data$predicted_activity)



ggplot(data = all_combined_cluster_2@meta.data) + 
  geom_histogram(mapping = aes(x = predicted_activity), color = "white", fill = "blue") +
  labs(x = 'Predicted Activity (%)', y = 'Nuclei', title = 'Rhemac10 snRNAseq cluster 2 Neuroestimator Results')


# save as rhemac10_snRNAseq_cluster_2_neuroestimator_results_hist.png




```



Can also subcluster the clusters in the predicted_activity violin plot that have a higher activity score and see where OSTN is expressed there








DIFFERENTIAL GENE EXPRESSION ANALYSIS

Pseudobulk vs Seurat FindMarkers

Look into how to do pseudobulk differential gene expression (DGE) in Seurat
https://satijalab.org/seurat/articles/de_vignette
AggregateExpression function https://satijalab.org/seurat/reference/aggregateexpression
- to use, need to use group.by = to specify how to group the cells to aggregate counts across the cells

Video tutorial https://www.youtube.com/watch?v=04gB2owLKus&t=79s

Goal is From entire snRNAseq dataset, take bottom 10% of cells and top 90% of cells and pseudobulk for DEG

Calculated What are the score ranges in the top/bottom 10% of all cells? in lines 2040-2065 above, but I just extracted neuroestimator scores from neuroestimator results csv

Would need to extract cell id's along with score in seurat object - would need entire row in seurat object metadata



Goal is to perform pseudobulk DE by grouping just two groups of cells:

Top 90% of cells by predicted_activity for ENTIRE snRNAseq dataset to start (all_combined seurat object) 

Bottom 10% of cells by predicted_activity for ENTIRE snRNAseq dataset to start (all_combined seurat object)

…and you don’t care about which sample the cells came from (i.e., treating the entire dataset as one pool since I combined all samples into one seurat object), can aggregate all "High" and all "Low" cells into just 2 pseudobulk samples.

```{r}


# Access neuroestimator scores from seurat object
scores <- all_combined$predicted_activity
scores

# Define thresholds
top_90 <- quantile(scores, 0.9, na.rm = TRUE)
top_90

bottom_10 <- quantile(scores, 0.1, na.rm = TRUE)
bottom_10

# Label cells by adding a pseudobulk_group column to seurat object
# initiate column. Fill with NA
all_combined$pseudobulk_group <- NA
# then populate with high or low
all_combined$pseudobulk_group[scores >= top_90] <- "High"
all_combined$pseudobulk_group[scores <= bottom_10] <- "Low"


# view seurat object metadata
View(all_combined@meta.data)

# pseudobulk group column has High if cell id is in top 90%, Low if cell id is in bottom 10%, or NA if neither!

```


Next, subset only high and low cells in a seurat object (run aggregate expression on this, NOT pseudo_counts?)
```{r}

# ne = neuroestimator 
ne_subset_all_combined <- subset(all_combined, subset = pseudobulk_group %in% c("High", "Low"))
ne_subset_all_combined 


# 23477 features across 27142 samples within 1 assay 
# Active assay: RNA (23477 features, 2000 variable features)

# ~27k nuclei in either top 90% ne scores or bottom 10% ne scores (REMINDER: total snRNAseq dataset has 135k nuclei)

```



Next, aggregate counts into two groups (high & low). May need to do again with fully annotated dataset (done on annotated dataset in tutorial here https://satijalab.org/seurat/articles/de_vignette)

Also follow youtube video here https://www.youtube.com/watch?v=04gB2owLKus&t=79s

Count aggregation to neuroestimator score level (high or low). The return.seurat needs to be TRUE
```{r}

# did originally
# pseudo_counts <- AggregateExpression(ne_subset_all_combined , assays = "RNA", return.seurat = T, group.by = "pseudobulk_group")
# seurat object only had 2 samples??? not right, redo below

# creating a list, and within list there is a counts matrix
# aggregated cells by high & low in pseudobulk_group column
# return a pseurat object (was false before)
pseudo_counts <- AggregateExpression(ne_subset_all_combined, assays = "RNA", return.seurat = TRUE, group.by = "pseudobulk_group", slot = "counts")
pseudo_counts


# each 'cell' is a neuroestimator high or low pseudobulk profile
tail(Cells(pseudo_counts))

# when return.seurat = FALSE (needs to be TRUE)
# $RNA
# 18691 x 2 sparse Matrix of class "dgCMatrix"
#                     High    Low
# ENSMMUG00000023296   315    526
# ZNF692               643   1219
# ZNF672               336    569
# SH3BP5L              553   1004
# ENSMMUG00000050812    18     37
# ENSMMUG00000064763    27     83
# OR11L1                18     42

# This aggregates raw counts into a pseudobulk-style matrix, where each column is the summed gene expression for "High" or "Low".



# when return.seurat = TRUE
# An object of class Seurat 
# 18691 features across 2 samples within 1 assay 
# Active assay: RNA (18691 features, 0 variable features)
#  3 layers present: counts, data, scale.data
# [1] "High" "Low" 

```



Retrieve counts data (data not like in youtube video)
```{r}

# pseudo_counts <- pseudo_counts$RNA
# View(pseudo_counts)

```






6/26/25

Find gene symbol or ensembl ID for the following genes: LINC00473 (PDE10A), LINC00507, LINC00602 & make feature plots if possible
LINC00473: PDE10A
LINC00507: only has human ensembl id
LINC00602: 

Go to gprofiler > g:orth > human to macaca mulatta
```{r}

FeaturePlot(all_combined, feature = "PDE10A")

```

Make full UMAP again but make numbers smaller to see if clusters easier to read (not much difference)
```{r}

# resolution 0.8
DimPlot(all_combined, reduction = "umap", label = TRUE, pt.size = NULL)
# 43 clusters
# save as png: GB_combined_rhemac10_snRNAseq_UMAP_PC1-30_res0.8_6_5_25.png


```



6/30/25

continue with pseudobulk video ~22:00? Using published 10x dataset with many sample names...
Data here is one dataset with 18 replicates combined...want to find DEG betwen high and low neuroestimator scores

doing sample-level differential expression analysis (tutorials so far are doing sample level dge), doing differential expression analysis based on neuroestimator scores from all 18 samples?

Already ran AggregateExpression() function in line 2671

Found another tutorial here: https://hbctraining.github.io/scRNA-seq/lessons/pseudobulk_DESeq2_scrnaseq.html
In tutorial aggregated matrix has cell types as rows and genes as columns (snRNAseq dataset isn't annotated yet)

```{r}

# review: pseudo_counts object is the aggregated counts matrix. Rows are genes expressed across ~27k nuclei in either top 90% ne scores or bottom 10% ne scores (REMINDER: total snRNAseq dataset has 135k nuclei)

# pseudo_counts is a sparse matrix of class dgCMatrix

# JK needed to be a Seurat object for it to work! Done above

```


Go back to Seurat tutorial https://satijalab.org/seurat/articles/de_vignette

"Next, we perform DE testing on the pseudobulk level for CD14 monocytes, and compare it against the previous single-cell-level DE results."

Here I'm doing Next, DE testing on the pseudobulk level for high vs low neuroestimator scores

Try below (got error because has to be a seurat object? No it's because there are no counts.....
```{r}


pseudobulk_ne <- FindMarkers(object = pseudo_counts, 
                         ident.1 = "High", 
                         ident.2 = "Low",
                         test.use = "DESeq2")
head(pseudobulk_ne, n = 15)

View(pseudo_counts@meta.data)
# has no counts & no metadata.......not what I want

	
# orig.ident
# pseudobulk_group
# High
# High
# High
# Low
# Low
# Low

```


View metadata of subsetted seurat object
```{r}

View(ne_subset_all_combined@meta.data)

```




Probably have to pseudobulk based on sample id's then & not just neuroestimator scores. It's in cell barcode name

When doing pseudobulk differential expression, you generally want to aggregate counts by sample (or biological replicate), not just by a numeric score like neuroestimator. This preserves biological variability and proper statistical inference.

Since your sample ID is encoded in the cell barcode, you’ll want to extract sample IDs from those barcodes first, then group cells by sample and condition (e.g., high vs low neuroestimator scores) before pseudobulking.

Extract the sample name for each row in your Seurat metadata
```{r}

# Assuming rownames(ne_subset_all_combined@meta.data) are the cell barcodes like above

# Extract the sample name part from barcode
# add column called sample_name
ne_subset_all_combined@meta.data$sample_name <- sapply(
  rownames(ne_subset_all_combined@meta.data), 
  function(x) {
    # Split by underscore
    parts <- unlist(strsplit(x, "_"))
    # sample name is everything except the last part (barcode + suffix)
    # So remove last 1 or 2 parts?... barcode is everything after the third underscore
    # example: sample_1MI1_so_AAACCCAAGTCCCGAC-1
    # parts = c("sample", "1MI1", "so", "AAACCCAAGTCCCGAC-1")
    # Sample name = paste first 3 parts
    paste(parts[1:3], collapse = "_")
  }
)

# Check new column
head(ne_subset_all_combined@meta.data$sample_name)


# print out unique sample names
unique(ne_subset_all_combined@meta.data$sample_name)

#  [1] "sample_1MI1_so" "sample_1MI3_so" "sample_2MI1_so" "sample_2MI3_so" "sample_3C_MI2"  "sample_3MI1_so" "sample_3MI3_so" "sample_4D_MI2"  "sample_4MI1_so" "sample_4MI3_so" "sample_5MI1_so"
# [12] "sample_5MI3_so" "sample_6MI1_so" "sample_6MI3_so" "sample_7MI1_so" "sample_7MI3_so" "sample_8MI1_so" "sample_8MI3_so"


```



View meta data now
```{r}

View(ne_subset_all_combined@meta.data)

```




7/2/25 cluster back up!

Continue pseudobulking

Use AggregateExpression() to pseudobulk based on sample id in sample id column as well as neuroestimator score (High or Low) in pseudobulk_group column of ne_subset_all_combined seurat object. Return a seurat object
```{r}

pseudo_counts_2 <- AggregateExpression(ne_subset_all_combined, assays = "RNA", return.seurat = TRUE, group.by = c("sample_name", "pseudobulk_group"), slot = "counts")
pseudo_counts_2



# each 'cell' is a sample_nescore- pseudobulk profile
# there are 18 samples. Each one has high or low label. 18 x 2 = 36
tail(Cells(pseudo_counts_2))


# > head(Cells(pseudo_counts_2))
# [1] "sample-1MI1-so_High" "sample-1MI1-so_Low"  "sample-1MI3-so_High" "sample-1MI3-so_Low"  "sample-2MI1-so_High" "sample-2MI1-so_Low" 


```



Next, we perform DE testing on the pseudobulk level high ne scores vs low ne scores
```{r}

pseudobulk_ne_2 <- FindMarkers(object = pseudo_counts_2, 
                         ident.1 = "High", 
                         ident.2 = "Low",
                         test.use = "DESeq2")
head(pseudobulk_ne_2, n = 15)

View(pseudo_counts_2@meta.data)


```

START HERE PSEUDOBULKING DGE

What happened...

When you use AggregateExpression(..., group.by = c("sample_name", "pseudobulk_group")), the new Seurat object pseudo_counts_2 has one cell (aggregate column) per combination of sample_name and pseudobulk_group. The cell names look like this: "sample-7MI3-so_High", "sample-7MI3-so_Low"

Fix identities for DE testing
To test "High" vs "Low" pseudobulk groups, you need to:

1) Set Idents to the pseudobulk_group only (i.e., extract High / Low from cell names).

2) Then run FindMarkers()


```{r}

# Step 1: Extract 'High' or 'Low' from the cell names
Idents(pseudo_counts_2) <- sapply(Cells(pseudo_counts_2), function(x) {
  sub(".*_", "", x)  # gets last part after the  underscore
})

# Check identities
table(Idents(pseudo_counts_2))
# Should show counts for 'High' and 'Low'



```


Install Deseq2 1st
```{r}
BiocManager::install("DESeq2")

```


Calculate DE genes between cells with high vs low neuroestimator scores for full dataset
```{r}
# need to install deseq2

# reminder: FindMarkers calculates DEG between 2 groups or two clusters

# Step 2: Run DESeq2-based DE analysis
pseudobulk_ne_2 <- FindMarkers(
  object = pseudo_counts_2, 
  ident.1 = "High", 
  ident.2 = "Low", 
  test.use = "DESeq2"
)

head(pseudobulk_ne_2, n = 15)

# pseudobulk_ne_2 <- FindMarkers(object = pseudo_counts_2, ident.1 = "High", ident.2 = "Low", test.use = "DESeq2")


# direction of comparison - cells with high ne scores vs cells with low ne scores - what genes are differentially expressed?

```



Worked!!!! Export as csv!
In cluster ~/macaque_snRNAseq make folder called pseudobulk_dge
```{r}

# subsetted all_combined (~135k nuclei) for nuclei that had highest and lowest ne scores (~27k) and pseudobulked them to calculate differentially expressed genes

write.csv(pseudobulk_ne_2, file = "~/macaque_snRNAseq/pseudobulk_dge/pseudobulk_dge_snRNAseq_high_vs_low_nescores_7_2_25.csv")

# export to umms computer using filezilla

# csv has 18,691 DEGs

# add to dropbox 
```



Calculate default Seurat differential gene expression test for ne_subset_all_combined seurat object (WITHOUT running AggregateExpression() function)
```{r}

# Find DE features between high and low ne scores

# temporarily change cell idents to pseudobulk_group metadata column. Change back to seurat_clusters later
Idents(ne_subset_all_combined) <- "pseudobulk_group"
ne_subset_all_combined_dge_default <- FindMarkers(ne_subset_all_combined, ident.1 = "High", ident.2 = "Low")
# view results
head(ne_subset_all_combined_dge_default)


```


Need to join layers (samples) 1st for subsetted seurat object. See lines 2234-2237
```{r}

ne_subset_all_combined_join <- JoinLayers(ne_subset_all_combined)
ne_subset_all_combined_join

```



Calculate default Seurat differential gene expression test for ne_subset_all_combined seurat object (WITHOUT psuedobulking & running AggregateExpression() function)
```{r}

# Find DE features between high and low ne scores

# temporarily change cell idents to pseudobulk_group metadata column. Change back to seurat_clusters later
Idents(ne_subset_all_combined_join) <- "pseudobulk_group"
ne_subset_all_combined_dge_default <- FindMarkers(ne_subset_all_combined_join, ident.1 = "High", ident.2 = "Low")
# view results
head(ne_subset_all_combined_dge_default)


```

Export results as CSV (don't save in pseudobulk_dge folder because NOT pseudobulking). Make a normal dge folder called dge or diff_gene_exp? Or just place csv in ~/macaque_snRNAseq/ for now

```{r}

write.csv(ne_subset_all_combined_dge_default, file = "~/macaque_snRNAseq/findmarkders_dge_snRNAseq_high_vs_low_nescores_default_output_7_2_25.csv")


```






7/9/25

SAVE entire dataset seurat object with pseudobulk_group column added in ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files
```{r}

saveRDS(all_combined, file = '~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_neuroestimator_pseudobulk_7_9_25.rds')


```



Also save seurat object subset of high & low ne scores from all_combineds
```{r}

saveRDS(ne_subset_all_combined, file = '~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/ne_subset_all_combined_7_9_25.rds')

```


Begin performing pseudobulking DGE on high vs low neuroestimator scores on JUST cluster 21 (DONE)

Review cluster 21 seurat object has less than 2k cells
```{r}

all_combined_cluster_21

```


Perform SAME steps as all_combined object but just for cluster 21 seurat object
```{r}

# Reference line 2616 above
# perform SAME steps as all_combined object but just for cluster 21 seurat object

# Access neuroestimator scores from seurat object
scores <- all_combined_cluster_21$predicted_activity
scores

# Define thresholds
top_90 <- quantile(scores, 0.9, na.rm = TRUE)
top_90

bottom_10 <- quantile(scores, 0.1, na.rm = TRUE)
bottom_10

# Label cells by adding a pseudobulk_group column to seurat object
# initiate column. Fill with NA
all_combined_cluster_21$pseudobulk_group <- NA
# then populate with high or low
all_combined_cluster_21$pseudobulk_group[scores >= top_90] <- "High"
all_combined_cluster_21$pseudobulk_group[scores <= bottom_10] <- "Low"


# view seurat object metadata
View(all_combined_cluster_21@meta.data)

# pseudobulk group column has High if cell id is in top 90%, Low if cell id is in bottom 10%, or NA if neither!


```



Next, subset only high and low cells in seurat object 

If too few cells, then will need to perform pseudobulking on cluster 2 AND 21
```{r}

# ne = neuroestimator 
ne_subset_all_combined_cluster21 <- subset(all_combined_cluster_21, subset = pseudobulk_group %in% c("High", "Low"))
ne_subset_all_combined_cluster21


# Warning: Removing 1456 cells missing data for vars requestedAn object of class Seurat 
# 23477 features across 366 samples within 1 assay 
# Active assay: RNA (23477 features, 2000 variable features)

# 336 nuclei in either top 90% ne scores or bottom 10% ne scores in cluster 21 (REMINDER: cluster 21 has 1822 nuclei)


```



Reference line 2819 Extract the sample name for each row in ne_subset_all_combined_cluster21
```{r}

# Extract the sample name part from barcode
# add column called sample_name
ne_subset_all_combined_cluster21@meta.data$sample_name <- sapply(
  rownames(ne_subset_all_combined_cluster21@meta.data), 
  function(x) {
    # Split by underscore
    parts <- unlist(strsplit(x, "_"))
    # sample name is everything except the last part (barcode + suffix)
    # So remove last 1 or 2 parts?... barcode is everything after the third underscore
    # example: sample_1MI1_so_AAACCCAAGTCCCGAC-1
    # parts = c("sample", "1MI1", "so", "AAACCCAAGTCCCGAC-1")
    # Sample name = paste first 3 parts
    paste(parts[1:3], collapse = "_")
  }
)


# check that sample_name column of ne_subset_all_combined_cluster21 contains 18 unique sample names (18 samples in full dataset) - yes!
# > length(unique(ne_subset_all_combined_cluster21@meta.data$sample_name))
# [1] 18

```



Refer to line 2869 Use AggregateExpression() to pseudobulk based on sample id in sample id column as well as neuroestimator score (High or Low) in pseudobulk_group column of ne_subset_all_combined seurat object. Return a seurat object
```{r}

pseudo_counts_21 <- AggregateExpression(ne_subset_all_combined_cluster21, assays = "RNA", return.seurat = TRUE, group.by = c("sample_name", "pseudobulk_group"), slot = "counts")
pseudo_counts_21


# Step 1: Extract 'High' or 'Low' from the cell names
Idents(pseudo_counts_21) <- sapply(Cells(pseudo_counts_21), function(x) {
  sub(".*_", "", x)  # gets last part after the  underscore
})

# Check identities
table(Idents(pseudo_counts_21))
# Should show counts for 'High' and 'Low'



```


Refer to line 2941 Calculate DE genes between cells with high vs low neuroestimator scores for full dataset
```{r}


# reminder: FindMarkers calculates DEG between 2 groups or two clusters

# Step 2: Run DESeq2-based DE analysis
pseudobulk_ne_21 <- FindMarkers(
  object = pseudo_counts_21, 
  ident.1 = "High", 
  ident.2 = "Low", 
  test.use = "DESeq2"
)

head(pseudobulk_ne_2, n = 15)


# pseudobulk_ne_21 <- FindMarkers(object = pseudo_counts_21, ident.1 = "High", ident.2 = "Low", test.use = "DESeq2")

# direction of comparison - cells with high ne scores vs cells with low ne scores - what genes are differentially expressed?

```


Export pseudobulk DGE results for cluster 21 as a csv file!
```{r}

# subsetted all_combined (~135k nuclei) for just cells in cluster 21. Extracted nuclei that had highest and lowest ne scores (~27k) and pseudobulked them to calculate differentially expressed genes

write.csv(pseudobulk_ne_21, file = "~/macaque_snRNAseq/pseudobulk_dge/pseudobulk_dge_snRNAseq_high_vs_low_nescores_cluster21_default_output_7_9_25.csv")

# export to umms computer using filezilla

# csv has 16,414 DEGs

# add to dropbox (can make excel sheet sorting log2FC & p_val_adj later)

```



Make a volcano plot of pseudobulk DEG using EnhancedVolcano R package!

Package documentation https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html

1st install package with bioconductor
Load along with greppel at top of script
```{r}

BiocManager::install('EnhancedVolcano')

```

Make volcano of pseudobulk DEG in cluster 21!
```{r}

# reduced point size, kept default labeling, added connectors. If you can make this version for all comparisons we do, that would be great. Thank you

# plot has black vertical lines  for log2FC. Has one horizontal black line for pvalue significance cutoff
# genes with log2FC greater than 1 (upregulated) or less than -1 (downregulated) are red (significant). Can be changed to 0.05



# Virtually all aspects of an EnhancedVolcano plot can be configured for the purposes of accommodating all types of statistical distributions and labelling preferences. By default, EnhancedVolcano will only attempt to label genes that pass the thresholds that you set for statistical significance, i.e., ‘pCutoff’ and ‘FCcutoff’. 



EnhancedVolcano(pseudobulk_ne_21,
    lab = rownames(pseudobulk_ne_21),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.01,
    FCcutoff = 1,
    title = 'Pseudobulk DGE Cluster 21 High vs Low Neuroestimator scores',
    col=c('black', 'black', 'black', 'red3'),
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE)

# expand and save plot as Pseudobulk_DGE_Cluster_21_High_vs_Low_Neuroestimator_scores_Volcano_7_9_25.png
# add plot to dropbox

```



7/10/25

Make volcano plot for full dataset pseudobulk DGE!
```{r}
# pseudobulk_ne_2 object created for full dataset in line 2954

EnhancedVolcano(pseudobulk_ne_2,
    lab = rownames(pseudobulk_ne_2),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.01,
    FCcutoff = 1,
    title = 'Pseudobulk DGE full Rhemac10 snRNAseq dataset High vs Low Neuroestimator scores',
    col=c('black', 'black', 'black', 'red3'),
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE)

# expand and save plot as Pseudobulk_DGE_full_rhemac10_dataset_High_vs_Low_Neuroestimator_scores_Volcano_7_10_25.png
# add plot to dropbox



```

Adjust cutoffs
```{r}

# pseudobulk_ne_2 object created for full dataset in line 2954

EnhancedVolcano(pseudobulk_ne_2,
    lab = rownames(pseudobulk_ne_2),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.05,
    FCcutoff = 3,
    title = 'Pseudobulk DGE full Rhemac10 snRNAseq dataset High vs Low Neuroestimator scores',
    col=c('black', 'black', 'black', 'red3'),
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE)


# is showing genes but too many labeled
# expand and save plot as Pseudobulk_DGE_full_rhemac10_dataset_High_vs_Low_Neuroestimator_scores_Volcano_7_10_25.png
# add plot to dropbox






```


labels left and right side - harder to see right side labels
```{r}



EnhancedVolcano(pseudobulk_ne_2,
    lab = rownames(pseudobulk_ne_2),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    title = 'Pseudobulk DGE full Rhemac10 snRNAseq dataset High vs Low Neuroestimator scores',
    col=c('black', 'black', 'black', 'red3'),
    pointSize = 2.0,
    labSize = 3.0,
    drawConnectors = TRUE,
    max.overlaps = 20)


# is showing genes but too many labeled
# expand and save plot as Pseudobulk_DGE_full_rhemac10_dataset_High_vs_Low_Neuroestimator_scores_Volcano_7_10_25.png
# add plot to dropbox




```


try decreasing label size from above
```{r}

EnhancedVolcano(pseudobulk_ne_2,
    lab = rownames(pseudobulk_ne_2),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    title = 'Pseudobulk DGE full Rhemac10 snRNAseq dataset High vs Low Neuroestimator scores',
    col=c('black', 'black', 'black', 'red3'),
    pointSize = 1.0,
    labSize = 2.5,
    drawConnectors = TRUE,
    max.overlaps = 20)


# is showing genes but too many labeled
# expand and save plot as Pseudobulk_DGE_full_rhemac10_dataset_High_vs_Low_Neuroestimator_scores_Volcano_7_10_25.png
# add plot to dropbox




```




try plotting top 10 genes with most significant p-adj value - top genes have lowest p-value
with EnhancedVolcano package no direct way to plot max number of genes?
can change one side of x-axis (right positive log2FC) but not left negative side?
```{r}

# Rank genes by adjusted p-value
top10 <- rownames(pseudobulk_ne_2[order(pseudobulk_ne_2$p_val_adj), ])[1:10]

EnhancedVolcano(pseudobulk_ne_2,
    lab = ifelse(rownames(pseudobulk_ne_2) %in% top10, rownames(pseudobulk_ne_2), ""),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    title = 'Top 10 DEGs: High vs Low Neuroestimator scores',
    col = c('black', 'black', 'black', 'red3'),
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE,
    max.overlaps = 100)



```

plot top 30 most significant genes by adjusted p-value - still not labeling the left side though
```{r}

# Rank genes by adjusted p-value
top10 <- rownames(pseudobulk_ne_2[order(pseudobulk_ne_2$p_val_adj), ])[1:30]

EnhancedVolcano(pseudobulk_ne_2,
    lab = ifelse(rownames(pseudobulk_ne_2) %in% top10, rownames(pseudobulk_ne_2), ""),
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = 0.05,
    FCcutoff = 1.5,
    title = 'Top 30 DEGs: High vs Low Neuroestimator scores',
    col = c('black', 'black', 'black', 'red3'),
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE,
    max.overlaps = 100)




```


try custom cutoffs for left and right sife of x-axis - not working properly
```{r}

# Create custom significance groups
pval_thresh <- 0.05
fc_left <- -1.0
fc_right <- 3.0

# Create custom color group labels
key_genes <- rownames(pseudobulk_ne_2)
sig <- with(pseudobulk_ne_2,
            ifelse(p_val_adj < pval_thresh & avg_log2FC < fc_left, "LFC_down",
            ifelse(p_val_adj < pval_thresh & avg_log2FC > fc_right, "LFC_up", "NS")))

# Choose colors
custom_colors <- c("NS" = "black", "LFC_down" = "red", "LFC_up" = "blue")

# Plot
EnhancedVolcano(pseudobulk_ne_2,
    lab = key_genes,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    pCutoff = pval_thresh,
    FCcutoff = 0,  # disable default FC cutoff
    title = 'Pseudobulk DGE full Rhemac10 snRNAseq dataset High vs Low Neuroestimator scores',
    colCustom = custom_colors[sig],
    pointSize = 3.0,
    labSize = 6.0,
    drawConnectors = TRUE,
    max.overlaps = 20)


```


7/25/25

Using the full dataset with 18 libraries (all_combined seurat object), make 3 separate objects, subcluster, and make feature plots for each
Excitatory neurons -> clusters 0, 2, 3, 4, 5, 6, 7, 8, 12, 13, 14, 15, 17, 19, 20, 21, 22, 23, 25, 26, 27, 29, 31, 35, 36, 38, 39, 42
Inhibitory neurons -> clusters 9, 10, 18, 24, 30, 34, 40
Non-neuronal cells -> 1, 11, 16, 28, 32, 33, 37, 41

subset
see how i did this above for cluster 2 lines 2448-2488


NOTE - have to re-run every time you start a new job on OOD
```{r}

# all_combined_excitatory <- subset(all_combined, idents = 0, 2, 3, 4, 5, 6, 7, 8, 12, 13, 14, 15, 17, 19, 20, 21, 22, 23, 25, 26, 27, 29, 31, 35, 36, 38, 39, 42)
# all_combined_inhibitory <- subset(all_combined, idents = 9, 10, 18, 24, 30, 34, 40)
# all_combined_non_neuronal <- subset(all_combined, idents = 1, 11, 16, 28, 32, 33, 37, 41)
# REMAKE all_combined_excitatory below
# all_combined_excitatory <- subset(all_combined, idents = c(0, 2, 3, 4, 5, 6, 7, 8, 12, 13, 14, 15, 17, 19, 20, 21, 22, 23, 25, 26, 27, 29, 31, 35, 36, 38, 39, 42))

all_combined_inhibitory <- subset(all_combined, idents = c(9, 10, 18, 24, 30, 34, 40))
# REMAKE all_combined_non_neuronal below
# all_combined_non_neuronal <- subset(all_combined, idents = c(1, 11, 16, 28, 32, 33, 37, 41))


```


See subsetted objects
```{r}

all_combined_excitatory
all_combined_inhibitory
all_combined_non_neuronal

```


Add cell totals
```{r}
100665+14301+20738

```



Subcluster
shortcut to pipe in r: command+shift+m
```{r}

all_combined_excitatory <- all_combined_excitatory %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)



all_combined_inhibitory <- all_combined_inhibitory %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)



all_combined_non_neuronal <- all_combined_non_neuronal %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)




```


Plot UMAPs
```{r}

DimPlot(all_combined_excitatory, reduction = "umap", label = TRUE)
DimPlot(all_combined_inhibitory, reduction = "umap", label = TRUE)
DimPlot(all_combined_non_neuronal, reduction = "umap", label = TRUE)

# save as png's

```


Save RDS in ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/ (done)
```{r}
saveRDS(all_combined_excitatory, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_excitatory.rds")
saveRDS(all_combined_inhibitory, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_inhibitory.rds")
saveRDS(all_combined_non_neuronal, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_non_neuronal.rds")

```



Read back in (when starting new job on ood)
ran 7/30/25
```{r}

all_combined_excitatory <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_excitatory.rds")
all_combined_inhibitory <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_inhibitory.rds")
all_combined_non_neuronal <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_non_neuronal.rds")

```



Feature plots I made for entire dataset above in lines 1517-1612


Feature plots mainly all_combined_excitatory & AD
```{r}


FeaturePlot(all_combined_excitatory, features = c("SLC17A7", "RORB", "FEZF2", "HPCAL1"))

FeaturePlot(all_combined_excitatory, features = c("THEMIS", "NXPH4", "OSTN", "NPY"))

FeaturePlot(all_combined_excitatory, features = c("PDE10A"))
# LINC00473 is PDE10A (activity-dependent)

# FeaturePlot(all_combined_excitatory, features = c("SST", "PVALB", "VIP", "LAMP5"))

FeaturePlot(all_combined_excitatory, features = c("FOS", "EGR1", "BDNF", "IGF1"))

FeaturePlot(all_combined_excitatory, features = c("NPAS4", "LINC00507", "ZNF331", "PER1"))
# LINC00507 isn't symbol in ensembl GTF. Would need to find mmul_10 ensembl id


# Cortical Excitatory Neurons
FeaturePlot(all_combined_excitatory, features = c("SATB2", "POU3F3", "POU3F2", "CUX2"))
FeaturePlot(all_combined_excitatory, features = c("TBR1", "ETV1", "BCL11B", "CUX1"))
FeaturePlot(all_combined_excitatory, features = c("FOXP1", "OTX1", "RELN"))


# # Cortical Inhibitory Neurons
# FeaturePlot(all_combined_excitatory, features = c("GAD2", "GAD1", "SLC32A1", "CALB1"))
# FeaturePlot(all_combined_excitatory, features = c("CALB2", "SST", "VIP", "PVALB"))
# FeaturePlot(all_combined_excitatory, features = c("GABRB1", "GABRB2", "GABRB3"))

FeaturePlot(all_combined_excitatory, features = c("FOXG1", "FOXP2", "MEF2C", "NPR3"))

FeaturePlot(all_combined_excitatory, features = "ARHGEF2")

```




Feature plots all_combined_inhibitory - mainly inhibitory marker genes & AD
```{r}

# FeaturePlot(all_combined_inhibitory, features = c("SLC17A7", "RORB", "FEZF2", "HPCAL1"))

# FeaturePlot(all_combined_inhibitory, features = c("THEMIS", "NXPH4", "OSTN", "NPY"))

FeaturePlot(all_combined_inhibitory, features = c("PDE10A", "LHX6", "NR2F2"))
# LINC00473 is PDE10A

FeaturePlot(all_combined_inhibitory, features = c("SST", "PVALB", "VIP", "LAMP5"))

FeaturePlot(all_combined_inhibitory, features = c("FOS", "EGR1", "BDNF", "IGF1"))
# activity dependent genes
# IFG1 activity-regulated gene specific to VIP neurons

FeaturePlot(all_combined_inhibitory, features = c("NPAS4", "LINC00507", "ZNF331", "PER1"))
# LINC00507 isn't symbol in ensembl GTF. Would need to find mmul_10 ensembl id
# more activity dependent genes


# # Cortical Excitatory Neurons
# FeaturePlot(all_combined_inhibitory, features = c("SATB2", "POU3F3", "POU3F2", "CUX2"))
# FeaturePlot(all_combined_inhibitory, features = c("TBR1", "ETV1", "BCL11B", "CUX1"))
# FeaturePlot(all_combined_inhibitory, features = c("FOXP1", "OTX1", "RELN"))


# Cortical Inhibitory Neurons
FeaturePlot(all_combined_inhibitory, features = c("GAD2", "GAD1", "SLC32A1", "CALB1"))
FeaturePlot(all_combined_inhibitory, features = c("CALB2", "SST", "VIP", "PVALB"))
FeaturePlot(all_combined_inhibitory, features = c("GABRB1", "GABRB2", "GABRB3"))


# FeaturePlot(all_combined_inhibitory, features = c("C1QB", "FLT1", "PDGFRA"))

FeaturePlot(all_combined_inhibitory, features = c("FOXG1", "FOXP2", "MEF2C", "NPR3"))

FeaturePlot(all_combined_inhibitory, features = "ARHGEF2")





```





Feature plots all_combined_non_neuronal - just non-neuronal marker genes
```{r}

FeaturePlot(all_combined_non_neuronal, features = c("GFAP", "S100B", "AQP4", "OLIG2"))

FeaturePlot(all_combined_non_neuronal, features = c("MBP", "ITGAM", "PTPRC", "C1QB"))

FeaturePlot(all_combined_non_neuronal, features = c("FLIT1", "PDGFRA", "PTPRC"))


```


7/28/25

	• TO DO: Use Seurats default Find Markers function to calculate differentially expressed genes in different predicted cell types
		○ Inhibitory clusters
			§ 30 vs 10
			§ 34 vs 9 & 10
			§ 40 vs 18
		○ Excitatory clusters
			§ 31 vs 4
			§ 39 vs 7
			§ 33 vs 11
			§ 37 vs 11
		○ Oligodendrocytes cluster
        16 vs 1

7/29/25

Put results in ~/macaque_snRNAseq/findmarkers_dge/

In dropbox make folder called findmarkers_dge: BoultingLab_shared / Experiments / snRNAseq Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF / findmarkers_dge

View combined object
```{r}

all_combined

```


Calculate differentially expressed genes in each cluster compared to ALL clusters and export as a csv
findAllMarkers ----------------- detects all differentially expressed markers for each cluster individually (cluster 0 vs all clusters, cluster 1 vs all clusters, etc)
use default function. Set to an object so you can use head function and save it as a csv file

findMarkers ----------------- detects all differentially expressed markers for one cluster or between 2 clusters




# use findMarkers() function to find markers between 2 clusters or between specific clusters (more than 2 clusters) in a seurat object

# use findAllMarkers() function to find markers in one cluster (or each cluster) compared to all other clusters in dataset

# find all markers docs seurat: https://satijalab.org/seurat/reference/findmarkers
# avg_logFC: log fold-chage of the average expression between the two groups. Positive values indicate that the gene is more highly expressed in the first group (first cluster)

IMPORTANT! Need to join sample layers (counts [raw counts], data [normalized counts], scale.data [scale.data] 1st before running any DGE)
Run on object where I joined layers (all_combined_join_layers)

Join layers
```{r}

# all_combined_inhibitory <- JoinLayers(all_combined_inhibitory)
# all_combined_excitatory <- JoinLayers(all_combined_excitatory)
# all_combined_non_neuronal <- JoinLayers(all_combined_non_neuronal)

```

Check that they are joined
```{r}
# all_combined_inhibitory
# all_combined_excitatory
# all_combined_non_neuronal
```

Made individual sub set seurat objects just to subcluster and create feature plots
When running FindMarkers, run on entire object but just specify idents/cluster numbers


Re-Read in entire dataset object & view
Located in ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_neuroestimator_pseudobulk_7_9_25.rds
```{r}

all_combined <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_neuroestimator_pseudobulk_7_9_25.rds")
all_combined

```

Join sample layers for entire dataset to perform DEG FindMarkers
```{r}

# join
all_combined_join_layers <- JoinLayers(all_combined)

# view object
all_combined_join_layers
# joined raw counts and normalized count layers (and scale.data layer)

```



Run FindMarkers on inhibitory clusters (done running) - 5-10 minutes per comparison
		○ Inhibitory clusters
			§ 30 vs 10
			§ 34 vs 9 & 10
			§ 40 vs 18
```{r}

deg_all_combined_30v10 <- FindMarkers(all_combined_join_layers, ident.1 = 30, ident.2 = 10)
head(deg_all_combined_30v10)
write.csv(deg_all_combined_30v10, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_30v10_072925.csv")



deg_all_combined_34v910 <- FindMarkers(all_combined_join_layers, ident.1 = 30, ident.2 = c(9,10))
head(deg_all_combined_34v910)
write.csv(deg_all_combined_34v910, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_34v910_072925.csv")



deg_all_combined_40v18 <- FindMarkers(all_combined_join_layers, ident.1 = 40, ident.2 = 18)
head(deg_all_combined_40v18)
write.csv(deg_all_combined_40v18, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_40v18_072925.csv")



```


Run FindMarkers Excitatory clusters (done)
		○ Excitatory clusters
			§ 31 vs 4
			§ 39 vs 7
			§ 33 vs 11
			§ 37 vs 11
```{r}

deg_all_combined_31v4 <- FindMarkers(all_combined_join_layers, ident.1 = 31, ident.2 = 4)
head(deg_all_combined_31v4)
write.csv(deg_all_combined_40v18, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_31v4_072925.csv")


deg_all_combined_39v7 <- FindMarkers(all_combined_join_layers, ident.1 = 39, ident.2 = 7)
head(deg_all_combined_39v7)
write.csv(deg_all_combined_39v7, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_39v7_072925.csv")


deg_all_combined_33v11 <- FindMarkers(all_combined_join_layers, ident.1 = 33, ident.2 = 11)
head(deg_all_combined_33v11)
write.csv(deg_all_combined_33v11, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_33v11_072925.csv")


deg_all_combined_37v11 <- FindMarkers(all_combined_join_layers, ident.1 = 37, ident.2 = 11)
head(deg_all_combined_37v11)
write.csv(deg_all_combined_37v11, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_37v11_072925.csv")


```




Run FindMarkers Oligodendrocytes cluster 
```{r}

deg_all_combined_16v1 <- FindMarkers(all_combined_join_layers, ident.1 = 16, ident.2 = 1)
head(deg_all_combined_16v1)
write.csv(deg_all_combined_16v1, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_16v1_072925.csv")


```


ALL 8 cluster comparisons are in cluster ~/macaque_snRNAseq/findmarkers_dge/

Use filezilla to copy files to local

Put result csv's in dropbox BoultingLab_shared / Experiments / snRNAseq Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF Mmul_10_snRNAseq_18_Libraries_MI1_MI2_MI3_Ensembl_FASTA_GTF/ findmarkers_dge



Added 8/8/25
Calculate DEG for cluster 21 vs cluster 6
```{r}

deg_all_combined_21v6 <- FindMarkers(all_combined_join_layers, ident.1 = 21, ident.2 = 6)
head(deg_all_combined_21v6)
write.csv(deg_all_combined_21v6, file = "~/macaque_snRNAseq/findmarkers_dge/deg_all_combined_21v6_072925.csv")

```



# Interpretation of FindMarkers output: 
# p_val - significance. If < 0.05 (5%) it's significant. Default is ordered by p-value
# avg_log2FC - average log2 fold change for every gene marker. If pos number it's up-regulated if neg number it's down regulated. The number listed is put in a power with base 2 (i.e. 2 ^6.9)
# pct.1 - % gene is detected in a speccific cluster
# pct.2 - % gene is detected in all clusters combined
# p_val_adj - adjusted p-value - Bonferroni correction is adjustment made to p-values when several dependent or independent statistical tests are performed simultaneously on a single data set
# comparing all DEG in ident.1 vs ident.2 (directionality of the comparison is left to right)

FindMarkers https://satijalab.org/seurat/reference/findmarkers


Refresher https://bioinformatics.ccr.cancer.gov/docs/getting-started-with-scrna-seq/Seurat_DifferentialExpression_Classification/



8/6/25

TO DO
Using all_combined_join_layers object...
Make cluster dendrogram (did this in GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd) (done)
Make dot plot using same genes as Wei et al Figure 1 (done)


Dot Plot Seurat https://satijalab.org/seurat/reference/dotplot
```{r}
# 20 genes total
wei_genes <- c("SLC17A7", "RORB", "FEZF2", "HPCAL1", "THEMIS", "NXPH4", "OSTN", "NPY", "GAD1", "LHX6", "NR2F2", "SST", "PVALB", "VIP", "LAMP5", "MOG", "PDGFRA", "AQP4", "FLT1", "C1QB")

DotPlot(object = all_combined_join_layers, features = wei_genes)

```

Make it red like paper
```{r}

DotPlot(object = all_combined_join_layers, features = wei_genes, cols = "red")

# printed out green?
# cols specifies specific color gradient - not a single color

```


Change color gradient
```{r}

DotPlot(object = all_combined_join_layers, features = wei_genes, cols = c("lightgrey", "red"))


```




Either expand window or screenshot or save using ggsave
```{r}


# save plot to an object
gene_dot_plot <- DotPlot(object = all_combined_join_layers, features = wei_genes, cols = c("lightgrey", "red"))


# save in cluster under where I saved all previous plots for figures with wei data: ~/macaque_snRNAseq/wei_2022_snRNAseq
# ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes.png", plot = gene_dot_plot, width = 12, height = 10, dpi = 300)


# genes still squished at bottom and image came out with black background? Resave (run in console)
# dimensions are much better but background is still black
ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes_2.png", plot = gene_dot_plot, width = 16, height = 12, dpi = 300)


# save using default coloring scheme
# background is still black
gene_dot_plot_default <- DotPlot(object = all_combined_join_layers, features = wei_genes)
ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes_3.png", plot = gene_dot_plot_default, width = 16, height = 12, dpi = 300)

```


Fix black background with ggsave
See github issue https://github.com/tidyverse/ggplot2/issues/4919
Fixed issue!
```{r}
# explicitly set a white background before saving the plot
gene_dot_plot <- gene_dot_plot + theme(
  panel.background = element_rect(fill = "white"),
  plot.background = element_rect(fill = "white"),
  legend.background = element_rect(fill = "white")
)

# save again
ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes_4.png", plot = gene_dot_plot, width = 16, height = 12, dpi = 300)
# add to dropbox ppt

```



Make cluster dendrogram (reference GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 3452-3552)

Build cluster tree function https://satijalab.org/seurat/reference/buildclustertree
Cluster tree function Seurat https://satijalab.org/seurat/reference/plotclustertree
```{r}

# PlotClusterTree(all_combined_join_layers)
# Warning: `PackageCheck()` was deprecated in SeuratObject 5.0.0.
# Please use `rlang::check_installed()` instead.Error: Cluster tree functionality requires 'ape' please install with 'install.packages('ape')'


# Solution I found!!!!!! https://github.com/satijalab/seurat/issues/1103
# boxed numbers are node labels, NOT cluster labels
# our dataset right now has 43 clusters
all_combined_join_layers <- BuildClusterTree(all_combined_join_layers)
# Error: Cluster tree functionality requires 'ape' please install with 'install.packages('ape')'
# library(ape)
cluster_dendrogram <- PlotClusterTree(all_combined_join_layers)
cluster_dendrogram


# screenshot and save as mmul10_full_snRNAseq_dataset_default_dendrogram.png
# add to dropbox



# understanding BuildClusterTree https://satijalab.org/seurat/reference/buildclustertree

# Constructs a phylogenetic tree relating the 'aggregate' cell from each identity class. Tree is estimated based on a distance matrix constructed in either gene expression space or PCA space.

# Note that the tree is calculated for an 'aggregate' cell, so gene expression or PC scores are summed across all cells in an identity class (cluster id) before the tree is constructed.



```

How to remove node numbers from dendrogram and make the cluster id's not start with g?
g is likely added during BuildClusterTree or PlotClusterTree profiles
Install ape package 1st!
Worked!
```{r}

# Extract the phylogenetic tree object
tree <- Tool(all_combined_join_layers, slot = "BuildClusterTree")

# Remove 'g' prefix if present in tip labels
tree$tip.label <- gsub("^g", "", tree$tip.label)

# Plot the tree manually without node numbers
# cex parameter character expansion in base R plotting functions
# cex = 1: Default text size.
# cex = 0.7: 70% of the default size (smaller).
# cex = 1.5: 150% of the default size (larger).
# leave cex ~0.7 (1 is too large)
plot.phylo(tree, type = "cladogram", show.node.label = FALSE, cex = 0.7)

# worked!

```

Make typical phylogenic tree/denfrogram with traditional straight lines
```{r}

plot.phylo(tree, type = "phylogram", show.node.label = FALSE, cex = 0.7)

# save as mmul10_full_snRNAseq_dataset_simple_dendrogram.png
# add to dropbox

```


Make direction horizontal
```{r}

plot.phylo(tree, type = "phylogram", show.node.label = FALSE, cex = 0.7, direction = "downwards")

# # save as mmul10_full_snRNAseq_dataset_simple_dendrogram_edit.png
# add to dropbox


```


How can I make the order of the clusters in the dengrogram the same on the dot plot and put it on the x-axis? Then have the genes as the y-axis (similar to Wei et al 2022 fig 1e)
```{r}

# Extract the order of cluster levels from the dendrogram
# cluster_order <- all_combined_join_layers@tools$BuildClusterTree$tree$tip.label
# cluster_order

# # Extract the cluster order from the tree tips in tree object
cluster_order <- tree$tip.label
cluster_order

# shows order from smallest to largest, not order in dendrogram


# This gives you the order of cluster labels as they appear in the dendrogram plot
dendrogram_order <- tree$tip.label[tree$order]
dendrogram_order




```

Easier to manually change order of clusters in dot plot and put on x-axis

1ST put cell identity on x-axis and gene name on y-axis (flip axes)
Try adding coord_flip() function https://github.com/satijalab/seurat/issues/4866
```{r}
# dot plot I want to use
# gene_dot_plot <- DotPlot(object = all_combined_join_layers, features = wei_genes, cols = c("lightgrey", "red"))


gene_dot_plot_flip <- DotPlot(object = all_combined_join_layers, features = wei_genes, cols = c("lightgrey", "red")) + coord_flip()
gene_dot_plot_flip

```



Great! Now just need to change cluster idents on x-axis

Check levels of object before I change order
```{r}

levels(Idents(all_combined_join_layers))

```



Vector of cluster id's that are the same as in the dendrogram
```{r}

cluster_order_dot_plot <- c(
  "29", "11", "37", "28", "36", "41", "1", "16", "9", "18",
  "24", "14", "6", "21", "27", "8", "39", "23", "42", "33",
  "19", "17", "5", "35", "13", "31", "4", "20", "0", "3",
  "22", "26", "32", "34", "40", "38", "15", "2", "12", "7",
  "25", "10", "30"
)


```


Make copy of seurat object for dotplot
```{r}
all_combined_join_layers_reorder_dotplot <- all_combined_join_layers

# view copied object
all_combined_join_layers_reorder_dotplot

```

Make dotplot with all_combined_join_layers_reorder_dotplot

check idents
```{r}

Idents(all_combined_join_layers_reorder_dotplot)
# Levels: 0 1 2 3 4 5 6 7 8 9 10 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25 26 27 28 29 30 31 32 33 34 35 36 37 38 39 40 41 42

```


Re-order cell idents then just make dot plot normally
```{r}

Idents(all_combined_join_layers_reorder_dotplot) <- factor(
  Idents(all_combined_join_layers_reorder_dotplot),
  levels = cluster_order_dot_plot
)

```



check that they are re-ordered
```{r}

Idents(all_combined_join_layers_reorder_dotplot)
# Levels: 29 11 37 28 36 41 1 16 9 18 24 14 6 21 27 8 39 23 42 33 19 17 5 35 13 31 4 20 0 3 22 26 32 34 40 38 15 2 12 7 25 10 30

```



Create custom order for x-axis
```{r}

gene_dot_plot_flip_edit <- DotPlot(object = all_combined_join_layers_reorder_dotplot, features = wei_genes, cols = c("lightgrey", "red")) + coord_flip()
gene_dot_plot_flip_edit


```

Save using ggsave
Make background white 1st (why black?)
```{r}

# explicitly set a white background before saving the plot
gene_dot_plot_flip_edit <- gene_dot_plot_flip_edit + theme(
  panel.background = element_rect(fill = "white"),
  plot.background = element_rect(fill = "white"),
  legend.background = element_rect(fill = "white")
)


# save
ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes_edit.png", plot = gene_dot_plot_flip_edit, width = 16, height = 12, dpi = 300)

# slack to GB 

```


edit dot plot w/ more genes 8/7/25
```{r}

wei_genes_add_2 <- c("SLC17A7", "RORB", "FEZF2", "HPCAL1", "THEMIS", "NXPH4", "OSTN", "NPY", "GAD1", "LHX6", "NR2F2", "SST", "PVALB", "VIP", "LAMP5", "MOG", "PDGFRA", "AQP4", "FLT1", "C1QB", "FSTL1", "EGR1", "NPAS4", "FOS", "PER1", "ZNF331", "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC")


gene_dot_plot_flip_edit_add_2 <- DotPlot(object = all_combined_join_layers_reorder_dotplot, features = wei_genes_add_2, cols = c("lightgrey", "red")) + coord_flip()
gene_dot_plot_flip_edit_add_2




```

save plot as svg - read into adobe illistrator
```{r}
# install.packages("svglite")
# load
library(svglite)

ggsave("~/macaque_snRNAseq/wei_2022_snRNAseq/dotplot_weigenes_edit_more_genes_2.svg", plot = gene_dot_plot_flip_edit_add_2, width = 16, height = 12, dpi = 300)

```






8/7/25

Make more feature plots for inhibitory neuron subcluster - made in line 3481. Reclustered, saved, made feature plots etc in lines 3521-3527, 3549, 3560, 3571, 3620-3657
```{r}

FeaturePlot(all_combined_inhibitory, features = c("ADARB2", "PCDH17", "CHRD", "TRPC3"))

FeaturePlot(all_combined_inhibitory, features = c("PAX6", "SYT10", "UNC5B", "CPLX2"))

FeaturePlot(all_combined_inhibitory, features = c("GPM6A", "PKIA", "IGFBP2", "SOCS2"))

FeaturePlot(all_combined_inhibitory, features = c("USP36", "LHX6", "RGS5"))

# FeaturePlot(all_combined_inhibitory, features = c("ADARB2", "PCDH17", "CHRD", "TRPC3"))



# save as all_combined_inhibitory_featureplots10.png, all_combined_inhibitory_featureplots11.png, all_combined_inhibitory_featureplots12.png, etc
# add png's to dropbox too in case needed for later

```




Continue making feature plots with full dataset UMAP on genes from Wei annotations that I didn't make feature plots of yet
Use genes from transfer_data_results_wei_onto_boulting_edited_7_24_25.csv that I haven't made feature plots for
Last made feature plots for full dataset in lines 1517-1612

```{r}

# excitatory
FeaturePlot(all_combined, features = c("CBLN4", "GPR83", "WFDC2", "CACNA1G"))

FeaturePlot(all_combined, features = c("SMARCA2", "FAM198B", "SMYD2", "HTR2C"))

FeaturePlot(all_combined, features = c("SEC11C", "SLC24A2", "ARC", "CD63"))

FeaturePlot(all_combined, features = c("CNTNAP5", "PHLDB2", "SNCB", "CD63"))

FeaturePlot(all_combined, features = c("SEC11C", "SLC24A2", "ARC", "TNK2"))


# inhibitory
FeaturePlot(all_combined, features = c("ADARB2", "PCDH17", "CHRD", "TRPC3"))

FeaturePlot(all_combined, features = c("PAX6", "SYT10", "UNC5B", "CPLX2"))

FeaturePlot(all_combined, features = c("GPM6A", "PKIA", "IGFBP2", "SOCS2"))

# added SLC18A3/VCHAT - not found because human gene name. Need to find macaque mulatta equivalent if any
# same name in macaca mulatta https://www.ncbi.nlm.nih.gov/datasets/gene/710053/
# found here https://useast.ensembl.org/Macaca_mulatta/Gene/Summary?db=core;g=ENSMMUG00000047299;r=9:47773657-47775255;t=ENSMMUT00000075608
FeaturePlot(all_combined, features = c("USP36", "RGS5", "SLC18A3"))


# Continue in GB_Mmul_10_snRNAseq_cellranger_count_ensembl_5_16_25.Rmd
# Name png's gb_mmul10_snRNAseq_featureplots_22.png, gb_mmul10_snRNAseq_featureplots_23.png, etc & add to dropbox umap feature plots folder 


# non-neuronal - already made feature plots with these
# microglia, oligo, OPC, vascular
# FeaturePlot(all_combined, features = c("C1QB", "MOG", "PDGFRA", "FLT1"))


```



Try making feature plot with ENSMMUG00000047299 (SLC18A3)
not found
```{r}

FeaturePlot(all_combined, features = "ENSMMUG00000047299")

FeaturePlot(all_combined, features = "SLC18A3")

```



RE-make excitatory neuron & non-neuornal subcluster based off manual annotations in combined_UMAP_PC1-30_res.0.8_cluster_manual_annotations_7_24_25.xlsx
Excitatory neuron clusters are 0, 2, 3, 4, 5, 6, 7, 8, 12, 13, 15, 17, 19, 20, 21, 22, 23, 25, 26, 27, 29, 31, 32, 33, 35, 38, 39, 42
Non-neuronal clusters are 1, 11, 14, 16, 28, 36, 37, & 41

```{r}

all_combined_excitatory <- subset(all_combined, idents = c(0, 2, 3, 4, 5, 6, 7, 8, 12, 13, 15, 17, 19, 20, 21, 22, 23, 25, 26, 27, 29, 31, 32, 33, 35, 38, 39, 42))

all_combined_non_neuronal <- subset(all_combined, idents = c(1, 11, 14, 16, 28, 36, 37, 41))

```


Re-subcluster
```{r}

all_combined_excitatory <- all_combined_excitatory %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)


all_combined_non_neuronal <- all_combined_non_neuronal %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)



```


Excitatory neuron & non-neuronal subcluster UMAP
```{r}

DimPlot(all_combined_excitatory, reduction = "umap", label = TRUE)
# save as all_combined_excitatory_umap_redo.png

DimPlot(all_combined_non_neuronal, reduction = "umap", label = TRUE)
# all_combined_non_neuronal_umap_redo.png 


```


Re-make feature plots for new excitatory subcluster & add to dropbox
```{r}

# # save as all_combined_excitatory_umap_redo_featureplots1.png, all_combined_excitatory_umap_redo_featureplots2.png, all_combined_excitatory_umap_redo_featureplots3.png, etc

FeaturePlot(all_combined_excitatory, features = c("SLC17A7", "RORB", "FEZF2", "HPCAL1"))

FeaturePlot(all_combined_excitatory, features = c("THEMIS", "NXPH4", "OSTN", "NPY"))


FeaturePlot(all_combined_excitatory, features = c("FOS", "EGR1", "BDNF", "IGF1"))

FeaturePlot(all_combined_excitatory, features = c("NPAS4", "LINC00507", "ZNF331", "PER1"))
# LINC00507 isn't symbol in ensembl GTF. Would need to find mmul_10 ensembl id


# Cortical Excitatory Neurons
FeaturePlot(all_combined_excitatory, features = c("SATB2", "POU3F3", "POU3F2", "CUX2"))
FeaturePlot(all_combined_excitatory, features = c("TBR1", "ETV1", "BCL11B", "CUX1"))
FeaturePlot(all_combined_excitatory, features = c("FOXP1", "OTX1", "RELN", "FOXG1"))


FeaturePlot(all_combined_excitatory, features = c("FOXP2", "MEF2C", "NPR3", "ARHGEF2"))

FeaturePlot(all_combined_excitatory, features = c("PDE10A", "CBLN4", "GPR83", "WFDC2"))
# LINC00473 is PDE10A (activity-dependent)


# excitatory
FeaturePlot(all_combined_excitatory, features = c("CBLN4", "GPR83", "WFDC2", "CACNA1G"))

FeaturePlot(all_combined_excitatory, features = c("SMARCA2", "FAM198B", "SMYD2", "HTR2C"))

FeaturePlot(all_combined_excitatory, features = c("SEC11C", "SLC24A2", "ARC", "CD63"))

FeaturePlot(all_combined_excitatory, features = c("CNTNAP5", "PHLDB2", "SNCB", "CD63"))

FeaturePlot(all_combined_excitatory, features = c("SEC11C", "SLC24A2", "ARC", "TNK2"))


# inhibitory
FeaturePlot(all_combined_excitatory, features = c("ADARB2", "PCDH17", "CHRD", "TRPC3"))

FeaturePlot(all_combined_excitatory, features = c("PAX6", "SYT10", "UNC5B", "CPLX2"))

FeaturePlot(all_combined_excitatory, features = c("GPM6A", "PKIA", "IGFBP2", "SOCS2"))

# added SLC18A3/VCHAT - not found because human gene name. Need to find macaque mulatta equivalent if any
# same name in macaca mulatta https://www.ncbi.nlm.nih.gov/datasets/gene/710053/
# found here https://useast.ensembl.org/Macaca_mulatta/Gene/Summary?db=core;g=ENSMMUG00000047299;r=9:47773657-47775255;t=ENSMMUT00000075608
FeaturePlot(all_combined_excitatory, features = c("USP36", "RGS5", "SLC18A3"))



```



Re-make feature plots for new non-neuronal subcluster & add to dropbox
```{r}
# name all_combined_non_neuronal_umap_redo_featureplots1.png, all_combined_non_neuronal_umap_redo_featureplots2.png, etc

FeaturePlot(all_combined_non_neuronal, features = c("GFAP", "S100B", "AQP4", "OLIG2"))

FeaturePlot(all_combined_non_neuronal, features = c("MBP", "ITGAM", "PTPRC", "C1QB"))

FeaturePlot(all_combined_non_neuronal, features = c("FLIT1", "PDGFRA", "PTPRC", "MOG"))

```

Put all umap feature plots in folder in dropbox!









8/14/25 note: cluster shut down aug 15th-18th.

Re-save rds files in cluster
All rds files for this script saved in ~/macaque_snRNAseq/ensembl_mmul10_so_rds_files


Delete the following old RDS files in cluster
all_combined_excitatory.rds
all_combined_inhibitory.rds
all_combined_non_neuronal.rds


Re-save seurat subcluster objects
saveRDS(all_combined, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_seurat_preprocess_6_6_25.rds")
```{r}

# save
saveRDS(all_combined_excitatory, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_excitatory_8_14_25.rds")
saveRDS(all_combined_inhibitory, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_inhibitory_8_14_25.rds")
saveRDS(all_combined_non_neuronal, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_non_neuronal_8_14_25.rds")


# read in again

# all_combined_excitatory <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_excitatory_8_14_25.rds")
# all_combined_inhibitory <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_inhibitory_8_14_25.rds")
# all_combined_non_neuronal <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_non_neuronal_8_14_25.rds")


```


Re-save full dataset seurat objects
```{r}
# save
# full dataset w/o sample layers joined
saveRDS(all_combined, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_8_14_25.rds")

# full dataset w/ sample layers joined
saveRDS(all_combined_join_layers, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_join_layers_8_14_25.rds")


# read in again

# all_combined <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_8_14_25.rds")
# all_combined_join_layers <- readRDS("~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_join_layers_8_14_25.rds")

```



Plots
Make UMAP less pixelated
https://github.com/satijalab/seurat/issues/3881

The rasterization is switched on by default if you have more than 50k points being plotted. You can switch it off by providing additional argument raster=FALSE
```{r}

unannot_umap <- DimPlot(all_combined, reduction = "umap", label = TRUE, raster=FALSE)
unannot_umap

```


Make non-annotated UMAP & save as svg file
make folder ensembl_mmul10_r_plots
Use FZ to scp
```{r}

unannot_umap <- DimPlot(all_combined, reduction = "umap", label = TRUE, raster=FALSE)

ggsave("~/macaque_snRNAseq/ensembl_mmul10_r_plots/rhemac10_snRNAseq_unannotated_umap_081425_edit2.svg", plot = unannot_umap, width = 16, height = 12, dpi = 300)


```



Make copy of seurat object to rename cluster id's
```{r}

all_combined_join_layers_annot <- all_combined_join_layers

# view copied object
all_combined_join_layers_annot

```



Rename clusters
Use combined_UMAP_PC1-30_res.0.8_cluster_manual_annotations_updated_7_24_25.xlsx using Wei labels
```{r}

# all_combined_join_layers_annot <- RenameIdents(
#   all_combined_join_layers_annot,
#   `0` = "Exc L1-3 RORB SMARCA2",
#   `1` = "Oligo L1-6 MOG",
#   `2` = "Cluster 2",
#   `3` = "Cluster 3",
#   `4` = "Cluster 4",
#   `5` = "Cluster 5",
#   `6` = "Cluster 6",
#   `7` = "Cluster 7",
#   `8` = "Cluster 8",
#   `9` = "Cluster 9",
#   `10` = "Cluster 10",
#   `11` = "Cluster 11",
#   `12` = "Cluster 12",
#   `13` = "Cluster 13",
#   `14` = "Cluster 14",
#   `15` = "Cluster 15",
#   `16` = "Cluster 16",
#   `17` = "Cluster 17",
#   `18` = "Cluster 18",
#   `19` = "Cluster 19",
#   `20` = "Cluster 20",
#   `21` = "Cluster 21",
#   `22` = "Cluster 22",
#   `23` = "Cluster 23",
#   `24` = "Cluster 24",
#   `25` = "Cluster 25",
#   `26` = "Cluster 26",
#   `27` = "Cluster 27",
#   `28` = "Cluster 28",
#   `29` = "Cluster 29",
#   `30` = "Cluster 30",
#   `31` = "Cluster 31",
#   `32` = "Cluster 32",
#   `33` = "Cluster 33",
#   `34` = "Cluster 34",
#   `35` = "Cluster 35",
#   `36` = "Cluster 36",
#   `37` = "Cluster 37",
#   `38` = "Cluster 38",
#   `39` = "Cluster 39",
#   `40` = "Cluster 40",
#   `41` = "Cluster 41",
#   `42` = "Cluster 42"
# )



all_combined_join_layers_annot <- RenameIdents(
  all_combined_join_layers_annot,
  `0`  = "Exc L1-3 RORB SMARCA2",
  `1`  = "Oligo L1-6 MOG",
  `2`  = "Exc L1-6 RORB FAM198B",
  `3`  = "Exc L1-3 RORB SMARCA2",
  `4`  = "Exc L1-3 HPCAL1 GPR83",
  `5`  = "Exc L4-6 FEZF2 SLC24A2",
  `6`  = "Exc L4-6 RORB CNTNAP5",
  `7`  = "Exc L4-6 FEZF2 HTR2C",
  `8`  = "Exc L4-6 FEZF2 SLC24A2",
  `9`  = "Inh L1-3 SST CPLX2",
  `10` = "Inh L1-3 PVALB UNC5B",
  `11` = "Astrocyte L1-6 AQP4",
  `12` = "Exc L1-6 RORB FAM198B",
  `13` = "Exc L1-6 RORB FAM198B",
  `14` = "OPC L1-6 PDGFRA",
  `15` = "Exc L1-6 RORB FAM198B",
  `16` = "Oligo L1-6 MOG",
  `17` = "Exc L4-6 FEZF2 SLC24A2",
  `18` = "Inh L1-3 VIP USP36",
  `19` = "Exc L1-3 HPCAL1 GPR83",
  `20` = "Exc L1-3 RORB SMARCA2",
  `21` = "Exc L4-6 RORB CNTNAP5",
  `22` = "Exc L1-3 RORB SMARCA2",
  `23` = "Exc L1-3 HPCAL1 GPR83",
  `24` = "Inh L1-3 LAMP5 TRPC3",
  `25` = "Exc L4-6 FEZF2 HTR2C",
  `26` = "Exc L4-6 FEZF2 SLC24A2",
  `27` = "Exc L4-6 FEZF2 HTR2C",
  `28` = "Microglia L1-6 C1QB",
  `29` = "Exc L1-3 RORB SMARCA2",
  `30` = "Inh L1-3 PVALB UNC5B",
  `31` = "Exc L1-3 HPCAL1 GPR83",
  `32` = "Exc L1-6 RORB FAM198B",
  `33` = "Exc L1-6 RORB FAM198B",
  `34` = "Inh L1-3 SST CPLX2",
  `35` = "Exc L4-6 FEZF2 SLC24A2",
  `36` = "Vascular L1-6 FLT1",
  `37` = "Astrocyte L1-6 AQP4",
  `38` = "Exc L1-6 RORB FAM198B",
  `39` = "Exc L4-6 FEZF2 SLC24A2",
  `40` = "Inh L1-3 VIP USP36",
  `41` = "Microglia L1-6 C1QB",
  `42` = "Exc L4-6 FEZF2 SLC24A2"
)


```



check idents
```{r}
# print current cluster ident names
print(table(Idents(all_combined_join_layers_annot)))

```

Grouped cells with same label together & doesn't keep original ident

Plot anyways and see
```{r}

unannot_umap <- DimPlot(all_combined_join_layers_annot, reduction = "umap", label = TRUE, raster=FALSE)
unannot_umap

```


Sort alphabetically by 1st letter
```{r}

# Sort the cluster names alphabetically
Idents(all_combined_join_layers_annot) <- factor(
  Idents(all_combined_join_layers_annot),
  levels = sort(unique(as.character(Idents(all_combined_join_layers_annot))))
)

unannot_umap <- DimPlot(all_combined_join_layers_annot, reduction = "umap", label = TRUE, raster=FALSE)
unannot_umap

```


Save as svg edit 3
double check with excel sheet that all cell types are there!Yes
```{r}

ggsave("~/macaque_snRNAseq/ensembl_mmul10_r_plots/rhemac10_snRNAseq_annotated_umap_081425_edit3.svg", plot = unannot_umap, width = 16, height = 12, dpi = 300)

```


Also save as pdf
```{r}

ggsave("~/macaque_snRNAseq/ensembl_mmul10_r_plots/rhemac10_snRNAseq_annotated_umap_081425.pdf", plot = unannot_umap, width = 16, height = 12, dpi = 300)


```


Edit dot plot
make copy of seurat object
```{r}

all_combined_join_layers_reorder_dotplot_2 <- all_combined_join_layers
all_combined_join_layers_reorder_dotplot_2
```




edit
```{r}
# Move OSTN up next to FSTL1
# FLT1, ITGAM, PTPRC, C1QB, PDGFRA, MOG, AQP4, GAD1, LAMP5, VIP, PLAVB, SST, NR2F2, LHX6, SLC17A7, HPCAL1, GPR83, RORB, CNTNAP5, FAM198B, SMARCA2, THEMIS, FEZF2, SLC24A2, HTR2C, NXPH4, OSTN, FSTL1, EGR1, "NPAS4", "FOS", "PER1", "ZNF331", "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC"

# reorder dotplot genes
# wei_genes_add_2 <- c("SLC17A7", "RORB", "FEZF2", "HPCAL1", "THEMIS", "NXPH4", "OSTN", "NPY", "GAD1", "LHX6", "NR2F2", "SST", "PVALB", "VIP", "LAMP5", "MOG", "PDGFRA", "AQP4", "FLT1", "C1QB", "FSTL1", "EGR1", "NPAS4", "FOS", "PER1", "ZNF331", "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC")


# genes
dotplot_genes_2 <- c(
  "FLT1", "ITGAM", "PTPRC", "C1QB", "PDGFRA", "MOG", "AQP4", 
  "GAD1", "LAMP5", "VIP", "PVALB", "SST", "NR2F2", "LHX6", 
  "SLC17A7", "HPCAL1", "GPR83", "RORB", "CNTNAP5", "FAM198B", 
  "SMARCA2", "THEMIS", "FEZF2", "SLC24A2", "HTR2C", "NXPH4", 
  "OSTN", "FSTL1", "EGR1", "NPAS4", "FOS", "PER1", "ZNF331", 
  "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC"
)


# change cluster order again
# 36, 28, 41, 14, 1, 16, 37, 11, 24, 18, 40, 10, 30, 9, 34, 4, 23, 31, 6, 21, 2, 12, 13, 15, 32, 33, 38, 0,3, 20, 22, 29, 5, 8, 17, 26, 35, 39, 42, 7, 25, 27

cluster_order_dot_plot_2 <- c(
  "36", "28", "41", "14", "1", "16", "37", "11", "24", "18", 
  "40", "10", "30", "9", "34", "4", "23", "31", "6", "21", 
  "2", "12", "13", "15", "32", "33", "38", "0", "3", "20", 
  "22", "29", "5", "8", "17", "26", "35", "39", "42", "7", 
  "25", "27"
)


# reorder cell idents
Idents(all_combined_join_layers_reorder_dotplot_2) <- factor( Idents(all_combined_join_layers_reorder_dotplot_2), levels = cluster_order_dot_plot_2)

# Ensure unique gene names for plotting
# rownames(all_combined_join_layers_reorder_dotplot) <- make.unique(rownames(all_combined_join_layers_reorder_dotplot))

# identify idents NA console
# Idents(all_combined_join_layers_reorder_dotplot_2) %>% anyNA()

# remove NA idents to fix issue below
# all_combined_join_layers_reorder_dotplot_2 <- subset(all_combined_join_layers_reorder_dotplot_2, cell_type %in% na.omit(seurat_obj$cell_type))

# taking a while...
# all_combined_join_layers_reorder_dotplot_2 <- subset(
#   all_combined_join_layers_reorder_dotplot_2,
#   idents = na.omit(Idents(all_combined_join_layers_reorder_dotplot_2))
# )

# Assign NA cells to "Unassigned" without removing them
Idents(all_combined_join_layers_reorder_dotplot_2)[is.na(Idents(all_combined_join_layers_reorder_dotplot_2))] <- "Unassigned"


gene_dot_plot_flip_edit_add_4 <- DotPlot(object = all_combined_join_layers_reorder_dotplot_2, features = dotplot_genes_2, cols = c("lightgrey", "red")) + coord_flip()

# Warning: non-unique values when setting 'row.names': Error in `.rowNamesDF<-`(x, value = value) : 
  # duplicate 'row.names' are not allowed


gene_dot_plot_flip_edit_add_4


# Warning: The following requested variables were not found: GPR83, FAM198B


```

no duplicate genes or rows



Save as svg
```{r}

ggsave("~/macaque_snRNAseq/ensembl_mmul10_r_plots/rhemac10_snRNAseq_editdotplot_081425.svg", plot = gene_dot_plot_flip_edit_add_3, width = 16, height = 12, dpi = 300)

```




8/19/25
Cluster back on! Copy added code from local script below


make new objects & re-try
PERIOOOOODDDDDDDD
restoring cluster 19
```{r}

all_combined_join_layers_reorder_dotplot1 <- all_combined_join_layers

dotplot_genes1 <- c(
  "FLT1", "ITGAM", "PTPRC", "C1QB", "PDGFRA", "MOG", "AQP4", 
  "GAD1", "LAMP5", "VIP", "PLAVB", "SST", "NR2F2", "LHX6", 
  "SLC17A7", "HPCAL1", "GPR83", "RORB", "CNTNAP5", "FAM198B", 
  "SMARCA2", "THEMIS", "FEZF2", "SLC24A2", "HTR2C", "NXPH4", 
  "OSTN", "FSTL1", "EGR1", "NPAS4", "FOS", "PER1", "ZNF331", 
  "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC"
)

# forgetting cluster 19! probs causing issues
cluster_order_dot_plot1 <- c(
  "36", "28", "41", "14", "1","16", "37", "11", "24", "18", "40", "10", "30", "9", "34", "4", "19", 
  "23", "31", "6", "21", "2", "12", "13", "15", "32", "33", "38", "0", "3", "20", "22", "29", 
  "5", "8", "17", "26", "35", "39", "42", "7", "25", "27"
)


Idents(all_combined_join_layers_reorder_dotplot1) <- factor(
  Idents(all_combined_join_layers_reorder_dotplot1),
  levels = cluster_order_dot_plot1
)


gene_dot_plot_flip_edit <- DotPlot(object = all_combined_join_layers_reorder_dotplot1, features = dotplot_genes, cols = c("lightgrey", "red")) + coord_flip()
gene_dot_plot_flip_edit




```

Save dot plot as svg in local
```{r}

ggsave("/Users/gb/rhemac10_snRNAseq_editdotplot2_081425.svg", plot = gene_dot_plot_flip_edit, width = 16, height = 12, dpi = 300)

```

Fix PVALB spelling
```{r}

all_combined_join_layers_reorder_dotplot2 <- all_combined_join_layers

dotplot_genes2 <- c(
  "FLT1", "ITGAM", "PTPRC", "C1QB", "PDGFRA", "MOG", "AQP4", 
  "GAD1", "LAMP5", "VIP", "PVALB", "SST", "NR2F2", "LHX6", 
  "SLC17A7", "HPCAL1", "GPR83", "RORB", "CNTNAP5", "FAM198B", 
  "SMARCA2", "THEMIS", "FEZF2", "SLC24A2", "HTR2C", "NXPH4", 
  "OSTN", "FSTL1", "EGR1", "NPAS4", "FOS", "PER1", "ZNF331", 
  "IGF1", "BDNF", "PDE10A", "HTR2A", "HTR1B", "ARC"
)

# forgetting cluster 19! probs causing issues
cluster_order_dot_plot2 <- c(
  "36", "28", "41", "14", "1","16", "37", "11", "24", "18", "40", "10", "30", "9", "34", "4", "19", 
  "23", "31", "6", "21", "2", "12", "13", "15", "32", "33", "38", "0", "3", "20", "22", "29", 
  "5", "8", "17", "26", "35", "39", "42", "7", "25", "27"
)


Idents(all_combined_join_layers_reorder_dotplot2) <- factor(
  Idents(all_combined_join_layers_reorder_dotplot2),
  levels = cluster_order_dot_plot2
)


gene_dot_plot_flip_edit2 <- DotPlot(object = all_combined_join_layers_reorder_dotplot2, features = dotplot_genes2, cols = c("lightgrey", "red")) + coord_flip()
gene_dot_plot_flip_edit2




```


Save dot plot as svg in local
```{r}

ggsave("/Users/gb/rhemac10_snRNAseq_editdotplot3_081425.svg", plot = gene_dot_plot_flip_edit2, width = 16, height = 12, dpi = 300)

```





Make UMAP of the 3 animals from the 18 libraries
Have sample name appended to cell barcode in all_combined so can easily add metadata column called animal/macaque (MI1, MI2, MI3)
```{r}

# Extract "MI1", "MI2", or "MI3" from the cell names and add to metadata
all_combined_join_layers$macaque <- str_extract(
  colnames(all_combined_join_layers),
  "MI[1-3]"
)

# Check
head(all_combined_join_layers$macaque)

# how many cells are from each animal sample?
table(all_combined_join_layers$macaque)
#   MI1   MI2   MI3 
# 64,677 15,051 55,976 




# Now plot UMAP by macaque
# make raster = FALSE when plotting more than 100kk cells it's set to TRUE automatically
animalumap <- DimPlot(all_combined_join_layers, group.by = "macaque", raster=FALSE)
animalumap

```

Save plot & add to dropbox
```{r}

ggsave("/Users/gb/rhemac10_snRNAseq_annotatedumapanimal_081425.svg", plot = animalumap, width = 16, height = 12, dpi = 300)

```



8/15/25

The annotation per animal is interesting. Could you somehow calculate the percentage of cells per cluster that are derived for each animal?

and the percentage of cells contributed by each of the animals to the data set as a whole?

% of cells from each animal for whole dataset
```{r}

# > all_combined_join_layers
# An object of class Seurat 
# 23477 features across 135704 samples within 1 assay 
# Active assay: RNA (23477 features, 2000 variable features)
#  3 layers present: counts, data, scale.data
#  2 dimensional reductions calculated: pca, umap
# > table(all_combined_join_layers$macaque)
# 
#   MI1   MI2   MI3 
# 64677 15051 55976 
# > 64677/135704
# [1] 0.4766035

# MI1 cells are 47.7% of whole dataset

# > 15051/135704
# [1] 0.1109105

# MI2 cells are 11.1% of whole dataset

# > 55976/135704
# [1] 0.412486

# MI3 is 41.2% of whole dataset



```


% of cells from each animal for each cluster in all_combined_join_layers object?
```{r}

library(dplyr)

# Create a table of counts: rows = clusters, columns = macaque from metadata column I added
cell_counts <- table(
  cluster = Idents(all_combined_join_layers),
  macaque = all_combined_join_layers$macaque
)


# > cell_counts
#        macaque
# cluster   MI1   MI2   MI3
#      0    331  2420 10721
#      1   5300   733  5731
#      2   6033  1560  3676
#      3  10240   651   292
#      4   5258  1288  3629
#      5   3623   859  3707
#      6   3409  1075  2288
#      7   3232   563  1970
#      8   2622   528  2293
#      9   2148   507  2338
#      10  2477   421  1996
#      11  2403   478  1992
#      12  2538   571  1083
#      13     0     0  3201
#      14  1217   330  1303
#      15  1474   423   807
#      16  1272   100  1215
#      17  1211   263  1102
#      18  1130   244  1046
#      19  1113   189   737
#      20  1499   335    17
#      21   980   233   609
#      22   536   166   710
#      23   525   211   476
#      24   574   143   418
#      25   528    97   436
#      26   455   136   305
#      27   181    33   528
#      28   402    69   252
#      29   580     2    25
#      30   283    56   182
#      31   266    58   195
#      32   101    43   132
#      33   142    44    71
#      34   123    26    95
#      35    82    24    87
#      36    97    51    39
#      37    51    40    79
#      38    76    27    57
#      39    90    17    23
#      40    41    20    33
#      41    19    14    55
#      42    15     3    25



# Convert to % within each cluster
cell_percent <- prop.table(cell_counts, margin = 1) * 100

# Turn into a data frame for easier viewing/plotting
cell_percent_df <- as.data.frame(cell_percent)

# View first few rows
head(cell_percent_df)


```


Export results as csv
```{r}

write.csv(cell_percent_df, file = "/Users/gb/rhemac10_snRNAseq_percentanimalpercluster_081525.csv")

```


Format for plotting with ggplot2
```{r}


# library(tidyr)
# library(ggplot2)
# 
# cell_percent_tidy <- as.data.frame(cell_percent_df) %>%
#   pivot_longer(cols = -cluster, names_to = "macaque", values_to = "Percent")
# 
# head(cell_percent_tidy)
# 
# 
# 
# ggplot(cell_percent_tidy, aes(x = Cluster, y = Percent, fill = Macaque)) +
#   geom_bar(stat = "identity") +
#   labs(y = "Percent of Cells", x = "Cluster") +
#   theme_minimal()


animal_plot <- ggplot(cell_percent_df, aes(x = cluster, 
                             y = Freq, 
                             fill = macaque)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  theme_minimal() +
  labs(title = "Percentage of Cells per Cluster by Macaque Animal",
       x = "Cluster", 
       y = "% of Cells",
       fill = "Macaque") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

animal_plot

```



Save as an svg 
```{r}

animal_plot <- ggplot(cell_percent_df, aes(x = cluster, 
                             y = Freq, 
                             fill = macaque)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  theme_minimal() +
  labs(title = "Percentage of Cells per Cluster by Macaque Animal",
       x = "Cluster", 
       y = "% of Cells",
       fill = "Macaque") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("/Users/gb/rhemac10_snRNAseq_animalbargraph_081525.svg", plot = animal_plot, width = 16, height = 12, dpi = 300)

```





8/21/25

From entire dataset object, subset out cluster 6 & 21
Will use cNMF python package to see how to predict activity of cell types
Github here https://github.com/dylkot/cNMF/

Input data: Input data can be provided in 3 ways:
-as a scanpy file ending in .h5ad containg counts as the data feature. See the PBMC dataset tutorial for an example of how to generate the Scanpy object from the data provided by 10X. Because Scanpy uses sparse matrices by default, the .h5ad data structure can take up much less memory than the raw counts matrix and can be much faster to load.
-as a raw tab-delimited text file containing row labels with cell IDs (barcodes) and column labels as gene IDs
-as a 10x-Genomics-formatted mtx directory. You provide the path to the counts.mtx file or counts.mtx.gz file to counts_fn. It expects there to be barcodes.tsv and genes.tsv in the directory as well

know which samples come from what cell barcodes, but what cluster do the cell barcodes come from? May need to run cnmf on just cluster 6 and 21 separatly


In cluster made folder for cmf output called ~/macaque_snRNAseq/cnmf


Subset
```{r}

all_combined_excitatory_6_21 <- subset(all_combined_join_layers, idents = c(6, 21))
all_combined_excitatory_6_21

# has ~8k cells

```

Extract raw counts matrix like done for cluster 21 above lines 2243-2257
```{r}
# arrow keyboard shortcut in R mac is opt + minus
all_combined_excitatory_6_21_raw_matrix <- LayerData(all_combined_excitatory_6_21, assay = "RNA", layer = "counts")
all_combined_excitatory_6_21_raw_matrix

# custom_colors in matrix? went away!

```



Save as rds file and convert to h5ad adata/anndata object for python 

See https://satijalab.org/seurat/archive/v2.4/conversion_vignette (no longer works...old version of seurat)

Convert function SeuratDisk https://mojaveazure.github.io/seurat-disk/reference/Convert.html
```{r}


# save as rds in /home/gb1-umw/macaque_snRNAseq/ensembl_mmul10_so_rds_files
saveRDS(all_combined_excitatory_6_21_raw_matrix, file = "~/macaque_snRNAseq/ensembl_mmul10_so_rds_files/all_combined_cluster6_21_count_matrix_8_22_25.rds")

# convert using SeuratDisk package
# all_combined_excitatory_6_21_raw_matrix_ad <- Convert(from = all_combined_excitatory_6_21_raw_matrix, to = 'anndata', filename = '~/macaque_snRNAseq/cnmf/all_combined_cluster6_21_count_matrix_8_22_25.h5ad')
# all_combined_excitatory_6_21_raw_matrix_ad

```



Try below https://www.reddit.com/r/bioinformatics/comments/1ht2qb8/converting_seurat_rds_to_h5ad/
```{r}

SeuratDisk::SaveH5Seurat(all_combined_excitatory_6_21_raw_matrix, "~/macaque_snRNAseq/all_combined_excitatory_6_21_raw_matrix.h5Seurat")

SeuratDisk::Convert("~/macaque_snRNAseq/all_combined_excitatory_6_21_raw_matrix.h5Seurat", dest = "h5ad")

```



Try using scCustomize package to convert from rds (more specifically dgCMatrix) into h5ad object https://github.com/samuel-marsh/scCustomize/issues/195
use as.anndata() function https://samuel-marsh.github.io/scCustomize/reference/as.anndata.html
```{r}

as.anndata(all_combined_excitatory_6_21_raw_matrix, file_path = "~/macaque_snRNAseq/cnmf/", file_name = "all_combined_cluster6_21_count_matrix_8_22_25.h5ad")

```



8/25/25
Instead of saving as RDS, save as Matrix Market format, which should easily be read into python? but need barcodes csv and genes csv? Yes
Use Matrix package to save all_combined_excitatory_6_21_raw_matrix as .mtx file


Read below files into python jupyter notebooks
```{r}
library(Matrix)

# Suppose your matrix is named mat (dgCMatrix)
writeMM(all_combined_excitatory_6_21_raw_matrix, file = "~/macaque_snRNAseq/all_combined_cluster6_21_count_matrix_8_25_25.mtx")

# save row (genes) and columns (rows)
writeLines(rownames(all_combined_excitatory_6_21_raw_matrix), "~/macaque_snRNAseq/all_combined_cluster6_21_count_matrix_features.csv")
writeLines(colnames(all_combined_excitatory_6_21_raw_matrix), "~/macaque_snRNAseq/all_combined_cluster6_21_count_matrix_barcodes.csv")


```









9/8/25

Met with Via Scientific about multiomics & snRNAseq analysis
-they will continue multiomics analysis once they have access to cluster
-8 V1 multiomics libraries
-8 ACC (area of brain used to study autism) multiomics libraries


Make QC violin plots for UMI (nCount_RNA), genes (nFeature_RNA), mitochondrial genes (percent.mt), and ribosomal genes (percent.rb) to check that the smallet cluster values don't look off


View all libraries seurat object metadata
```{r}

View(all_combined@meta.data)


```



UMI violin plot
```{r}

# Visualize QC metrics as a violin plot
VlnPlot(all_combined, features = 'nFeature_RNA', pt.size = FALSE)

VlnPlot(all_combined, features = 'nCount_RNA', pt.size = FALSE)

VlnPlot(all_combined, features = 'percent.mt', pt.size = FALSE)

VlnPlot(all_combined, features = 'percent.rb', pt.size = FALSE)

```


Try exanding plot dimensions by saving using ggsave
```{r}

# Visualize QC metrics as a violin plot
qc_violin_genes <- VlnPlot(all_combined, features = 'nFeature_RNA', pt.size = FALSE)

qc_violin_umi <- VlnPlot(all_combined, features = 'nCount_RNA', pt.size = FALSE)

qc_violin_mt <- VlnPlot(all_combined, features = 'percent.mt', pt.size = FALSE)

qc_violin_rb <- VlnPlot(all_combined, features = 'percent.rb', pt.size = FALSE)



# save using ggsave
ggsave("/home/gb1-umw/macaque_snRNAseq/figures/qc_violin_genes.png", plot = qc_violin_genes, width = 16, height = 12, dpi = 300)

ggsave("/home/gb1-umw/macaque_snRNAseq/figures/qc_violin_umi.png", plot = qc_violin_umi, width = 16, height = 12, dpi = 300)

ggsave("/home/gb1-umw/macaque_snRNAseq/figures/qc_violin_mt.png", plot = qc_violin_mt, width = 16, height = 12, dpi = 300)

ggsave("/home/gb1-umw/macaque_snRNAseq/figures/qc_violin_rb.png", plot = qc_violin_rb, width = 16, height = 12, dpi = 300)


```




9/11/25

Subcluster clusters 6 & 21 to create UMAP

object
```{r}
all_combined_excitatory_6_21

# view metadata
View(all_combined_excitatory_6_21@meta.data)
```

Add metadata column for object to specify what animal the cell id comes from
Did before for entire R object above in lines 
```{r}

# Extract "MI1", "MI2", or "MI3" from the cell names and add to metadata
all_combined_excitatory_6_21$macaque <- str_extract(
  colnames(all_combined_excitatory_6_21),
  "MI[1-3]"
)

# Check
head(all_combined_excitatory_6_21$macaque)

# view metadata
View(all_combined_excitatory_6_21@meta.data)

```

Perform subclustering
Reference above lines 4306-4328
Reference GB_rhemac10_3'_snRNAseq_workflow_cellranger_2_21_24.Rmd lines 3979-4112
```{r}
# all_combined_excitatory_6_21 <- all_combined_excitatory_6_21 %>% 
#   NormalizeData() %>%
#   FindVariableFeatures() %>%
#   ScaleData() %>%
#   RunPCA() %>%
#   FindNeighbors(dims = 1:30) %>%
#   FindClusters() %>%
#   RunUMAP(dims = 1:30)


```

2000 variable feature calculated, run other steps
```{r}
# all_combined_excitatory_6_21 <- all_combined_excitatory_6_21 %>%
#   ScaleData() %>%
#   RunPCA() %>%
#   FindNeighbors(dims = 1:30) %>%
#   FindClusters() %>%
#   RunUMAP(dims = 1:30)

```



DON'T subset from seurat object that has layers joined - causes errors when subclustering
When you subset a Seurat object using subset(), sometimes cell metadata or dimension reduction slots (pca, umap) carry over references to the full object. Then, when you try to ScaleData() or RunPCA(), Seurat tries to write new data for the subset but conflicts with the old slots, giving: Cannot add new cells with [[<-
```{r}


# # make copy of subset
# all_combined_excitatory_6_21_reclust <- subset(all_combined_join_layers, idents = c(6, 21))
# 
# 
# # Drop old variable features/PCA/UMAP from the subset
# VariableFeatures(all_combined_excitatory_6_21_reclust) <- NULL
# all_combined_excitatory_6_21_reclust[["pca"]] <- NULL
# all_combined_excitatory_6_21_reclust[["umap"]] <- NULL
# 
# all_combined_excitatory_6_21_reclust <- FindVariableFeatures(all_combined_excitatory_6_21_reclust)
# 
# # 3. Scale the data
# all_combined_excitatory_6_21_reclust <- ScaleData(all_combined_excitatory_6_21_reclust, verbose = TRUE)
# 
# # 4. Run PCA
# all_combined_excitatory_6_21_reclust <- RunPCA(all_combined_excitatory_6_21_reclust, verbose = TRUE)
# 
# # 5. Find neighbors
# all_combined_excitatory_6_21_reclust <- FindNeighbors(all_combined_excitatory_6_21_reclust, dims = 1:30)
# 
# # 6. Find clusters
# all_combined_excitatory_6_21_reclust <- FindClusters(all_combined_excitatory_6_21_reclust)
# 
# # 7. Run UMAP
# all_combined_excitatory_6_21_reclust <- RunUMAP(all_combined_excitatory_6_21_reclust, dims = 1:30)


```



When subsetted above lines 4299 & 4301, subsetted from all_combined object, NOT the all_combined_join_layers object!
I think it is indeed related to #7804, if you try to "split" or "JoinLayers" after "subset", either way there is going to be errors!! https://github.com/satijalab/seurat/issues/8407
```{r}

# make copy of subset
all_combined_excitatory_6_21_reclust <- subset(all_combined, idents = c(6, 21))
all_combined_excitatory_6_21_reclust


```


re-cluster
```{r}

all_combined_excitatory_6_21_reclust <- all_combined_excitatory_6_21_reclust %>% 
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  FindNeighbors(dims = 1:30) %>%
  FindClusters() %>%
  RunUMAP(dims = 1:30)

```


Plot umap
```{r}

DimPlot(all_combined_excitatory_6_21_reclust, reduction = "umap", label = TRUE)
# save as rhemac10_snRNAseq_recluster_cluster6_21_9_10_25.png and add to dropbox

```

Add metadata
```{r}
# Extract "MI1", "MI2", or "MI3" from the cell names and add to metadata
all_combined_excitatory_6_21_reclust$macaque <- str_extract(
  colnames(all_combined_excitatory_6_21_reclust),
  "MI[1-3]"
)

# view metadata - scroll to last column in viewer
View(all_combined_excitatory_6_21_reclust@meta.data)

```


Color umap by macaque (MI1, MI2, or MI3) to see if clustering is driven by differences in cell expressio and not by technical artifacts (sample)
```{r}

reclust_animalumap <- DimPlot(all_combined_excitatory_6_21_reclust, group.by = "macaque", raster=FALSE)

reclust_animalumap
# save as rhemac10_snRNAseq_recluster_cluster6_21_macaque_9_10_25.png and add to dropbox

# out of the 18 libraries/samples, 2 are from MI2, 8 are from MI1, and 8 are from MI3
# cluster 6 & 21 split into 14 clusters 
# cluster 0 mainly seems to be from MI1, cluster 6 mainly from MI3
# cluster 1, 9, 5, 3 10, 2, 11 mainly from MI1
# clusters 12 and 8 from MI3

```

Add macaque metadta column to all_combined_excitatory_6_21_reclust BEFORE making umap




Make feature plot of OSTN
```{r}

FeaturePlot(all_combined_excitatory_6_21_reclust, features = 'OSTN')
# save as rhemac10_snRNAseq_recluster_cluster6_21_OSTN_9_10_25.png

```






9/15/25

color subclustered umap by either cluster 6 or cluster 21 to see representation of clustered cells
view metadata of seurat object all_combined_excitatory_6_21_reclust to color by metadata - need to add a metadata column
```{r}
View(all_combined_excitatory_6_21_reclust@meta.data)
```



Add metadata column called orig_cluster
```{r}
# Add original cluster identity as metadata
# The seurat_clusters column has the cluster id's
# all_combined_excitatory_6_21_reclust was created with line 5319 above
# Cells(all_combined_excitatory_6_21_reclust) should return the vector of cell barcodes in reclustered object
# all_combined$seurat_clusters[...] looks up the seurat_clusters metadata column in the original object (all_combined)

all_combined_excitatory_6_21_reclust$orig_cluster <- all_combined$seurat_clusters[
  Cells(all_combined_excitatory_6_21_reclust)
]

# Plot UMAP colored by original cluster (6 vs 21)
DimPlot(
  all_combined_excitatory_6_21_reclust,
  reduction = "umap",
  group.by = "orig_cluster",
  label = TRUE
)

# save as rhemac10_snRNAseq_recluster_cluster6_21_colorcluster_9_15_25.png
# results make sense - cluster 21 had the highest level of OSTN expression and it matches with the OSTN feature plot

```






9/17/25

TO DO
For each cluster in the dataset so far, perfrom pseudobulk DGE (already did for just cluster 21 - see above for reminder)

How I performed pseudobulking for cluster 21 above in lines 3064-3212


Performed pseudobulking on entire dataset (all clusters high vs low neuroestimator scores) in lines: 2622-2905 (did at 1st but didn't work); 2910-2978 (redid and worked)
Performed pseudo bulking for cluster 21 in lines: 3064-3212
	□ Created all_combined_cluster_21 seurat object in line 2224
Started doing pseudobulking in line 2594
	□ Goal is to perform pseudobulk DE by grouping just two groups of cells:
	□ Top 90% of cells by predicted_activity for ENTIRE snRNAseq dataset to start (all_combined seurat object) 
	□ Bottom 10% of cells by predicted_activity for ENTIRE snRNAseq dataset to start (all_combined seurat object)
	□ And you don’t care about which sample the cells came from (i.e., treating the entire dataset as one pool since I combined all samples into one seurat object), can aggregate all "High" and all "Low" cells into just 2 pseudobulk samples.
	
REMINDER: Need to perform DGE tests between two groups (cluster1 vs cluster2, control vs stimulated, etc). So when I pseudo bulked, I did it for cells with the highest neuroestimator score vs lower neuroestimator score

View seurat object for entire dataset
```{r}

all_combined

```

STEP 1
start by performing pseudobulking on cluster 6 (2nd highest expression of OSTN)
subset out cluster 6 form all_combined
```{r}

all_combined_cluster6 <- subset(all_combined, idents = 6)
all_combined_cluster6

```


STEP 2
view metadata to make sure it has predicted_activity score (neuroestimator score)
also already has the pseudobulk_group metadata column
```{r}

View(all_combined_cluster6@meta.data)

```


STEP 3
create object for the top & bottom neuroestimator scores
in this object there is only 1,148 nuclei because these are the nuclei that have a high ne score or low ne score
The all_combined_cluster6 object is a subset with ALL nuclei in cluster 6 (includes cells that have a high neuroestimator score, a low neuroestimator score)
```{r}

ne_subset_all_combined_cluster6 <- subset(all_combined_cluster6, subset = pseudobulk_group %in% c("High", "Low"))
ne_subset_all_combined_cluster6

```


STEP 4
Extract sample names from cell barcodes in ne object ne_subset_all_combined_cluster6
what this is doing is adding a sample_name metadata column
```{r}

# Extract the sample name part from barcode
# add column called sample_name
# use sapply to go through each cell barcode row

ne_subset_all_combined_cluster6@meta.data$sample_name <- sapply(
	rownames(ne_subset_all_combined_cluster6@meta.data), 
		function(x) {
		# Split by underscore
		  parts <- unlist(strsplit(x, "_"))
			# sample name is everything except the last part (barcode + suffix)
			# So remove last 1 or 2 parts?... barcode is everything after the third underscore
			# example: sample_1MI1_so_AAACCCAAGTCCCGAC-1
			# parts = c("sample", "1MI1", "so", "AAACCCAAGTCCCGAC-1")
			# Sample name = paste first 3 parts
			paste(parts[1:3], collapse = "_")
		}
)

```


STEP 5
View metadata of ne_subset_all_combined_cluster6
```{r}

head(ne_subset_all_combined_cluster6@meta.data$sample_name)

```


STEP 6
Run pseudobulking function AggregateExpression based on sample names (18 samples total)
Tutorial here https://satijalab.org/seurat/articles/de_vignette
```{r}
pseudo_counts_6 <- AggregateExpression(ne_subset_all_combined_cluster6, assays = "RNA", return.seurat = TRUE, group.by = c("sample_name", "pseudobulk_group"), slot = "counts")
pseudo_counts_6

```

STEP 7
Extract high vs low from cell id's for cluster 6
```{r}
# Extract 'High' or 'Low' from the cell names
Idents(pseudo_counts_6) <- sapply(Cells(pseudo_counts_6), function(x) {
  sub(".*_", "", x)  # get last part after underscore
})

# Check identities
table(Idents(pseudo_counts_6))


```

STEP 8
Next, we perform DE testing on the pseudobulk level for cells with high ne score in cluster 6 vs cells w/ low ne score in cluster 6, and compare it against the previous single-cell-level DE results.
```{r}
# step 2: Run DESeq2-Based DGE analysis

pseudobulk_ne_6 <- FindMarkers(
  object = pseudo_counts_6,
  ident.1 = "High",
  ident.2 = "Low",
  test.use = "DESeq2"
)

# show top 15 DEG using pseudobulk analysis
head(pseudobulk_ne_6, n = 15)

# LINC00473 (PDE10A) is at the top!

```


STEP 9
Save results as a csv file
```{r}

write.csv(pseudobulk_ne_6, file = "~/macaque_snRNAseq/pseudobulk_dge/pseudobulk_dge_snRNAseq_high_vs_low_nescores_cluster6_default_output_9_18_25.csv")

# reminder: default dge automatically outputs DEGs from highest to lowest adjusted p-value (significance). The lower the adjusted p-value, the more significantly expressed the gene is
# a positive log2FC means that the gene us upregulated in cells with high vs low ne scores
# a negative log2 FC means that gene is downregulated in cells with high vs low ne scores
# if you want to see just fold change and not just log2FC, need to calculate 2^number listed 

```



See how to run rest of clusters in parallel or in a loop/function so don't have to manually run for each cluster?
```{r}




```










9/22/25
From looking at cnmf results, see other ad genes expressed in clusters 6 & 21
```{r}

FeaturePlot(all_combined_excitatory_6_21_reclust, features = c('FSTL1', 'PCSK1', 'EGR1', 'HTR2A', 'PDE10A'))

# export as a png file with dimensions so more clear and don't have to wait for it to unsmush

```


Export as png
```{r}

cnmf_featureplots_usage1_genes <- FeaturePlot(all_combined_excitatory_6_21_reclust, features = c('FSTL1', 'PCSK1', 'EGR1', 'HTR2A', 'PDE10A'))


ggsave("/home/gb1-umw/macaque_snRNAseq/figures/cnmf_usage1_feature_plots_9_22_25.png", plot = cnmf_featureplots_usage1_genes, width = 16, height = 12, dpi = 300)


```



Make feature plots of more genes from Ataman et al 2016

```{r}

cnmf_featureplots_usage1_genes2 <- FeaturePlot(all_combined_excitatory_6_21_reclust, features = c('BDNF', 'GPR22', 'ADCYAP1'))

ggsave("/home/gb1-umw/macaque_snRNAseq/figures/cnmf_usage1_feature_plots2_9_22_25.png", plot = cnmf_featureplots_usage1_genes2, width = 16, height = 12, dpi = 300)

```



Pausing snRNAseq analysis for a bit to focus on merfish & multiomics analysis!








